---
title: "SOC-5811 Week 5: Sampling distributions and statistical inference"
author: "Nick Graetz"
date: "9/29/2025"
output: 
    beamer_presentation:
        incremental: false
        includes: 
            in_header: Minnesota.tex
---

```{r setup, include=FALSE}
## Libraries
library(tidyverse)
library(data.table)
library(ggplot2)
library(haven)
library(formatR)
library(marginaleffects)
library(cowplot)
# knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 20), tidy = TRUE)
knitr::opts_chunk$set(
 fig.width = 6,
 fig.height = 4,
 warning = FALSE, 
 message = FALSE
)
```

---

\frametitle{Review}

\textbf{Why are we creating models (e.g., linear regression) for data at all?}

* Summarizing relationships.

* Making predictions.

---

\frametitle{Review}

1. Choose a functional form for the model.

$$pcthouse_i=\beta_0+\beta_1{pctpop_i}+\epsilon_i$$

\textbf{Once we define a model, how do we estimate the parameters using data?}

---

\frametitle{Review}

1. Choose a functional form for the model.

$$pcthouse_i=\beta_0+\beta_1{pctpop_i}+\epsilon_i$$

2. Choose an estimator (i.e., objective function) that relates the model to real observed data (e.g., OLS).

$$\text{Sum of squared residuals}=\sum_{i=1}^n (y_i-\widehat{\beta_0}-\widehat{\beta_1}x_i)^2$$

3. Fit the model (e.g., estimate parameters) by optimizing the objective function.

---

\frametitle{Review}

\textbf{What is happening behind the scenes to fit the model?}

* If there is a closed-form solution: math.
* If there is not a closed-form solution: iterative calculations and approximations.

\begin{center}
\includegraphics[width = 0.7\textwidth]{images/ss_example.png}
\end{center}

---

\frametitle{Review}

* Linear regression is practically very useful because it summarizes relationships with simple parameters that are easy to interpret. 

* This is true even once we start adding _more_ independent variables!

---

\frametitle{Statistical inference}

* Across working years, are incomes higher at older ages in Hennepin County?

* If we can agree on definitions of income and age, then this seems like a question we should be able to answer empirically!

---

\frametitle{Statistical inference}

\tiny
```{r, echo=F}
msp_micro <- fread('https://raw.githubusercontent.com/ngraetz/soc5811/refs/heads/main/data/acs2023_hennepin.csv')
ggplot(data=msp_micro[educ=='4 years of college' & age %in% 25:64 & !is.na(hhincome),],
       aes(x=age,y=hhincome)) + 
  geom_point(alpha=0.7) + 
  scale_fill_viridis_c(option='plasma') + 
  labs(x='Age', y='Household income') + 
  theme_bw()
```

---

\frametitle{Statistical inference}

\tiny
```{r, echo=F}
ggplot(data=msp_micro[educ=='4 years of college' & age %in% 25:64 & !is.na(hhincome),],
       aes(x=age,y=hhincome)) + 
  geom_hex(alpha=0.7) + 
  scale_fill_viridis_c(option='plasma') + 
  labs(x='Age', y='Household income') +
  theme_bw()
```

---

\frametitle{Statistical inference}

\tiny
```{r, echo=F}
msp_micro_working <- msp_micro[educ=='4 years of college' & age %in% 25:64 & !is.na(hhincome),]
msp_micro_working[, age_factor := as.factor(age)]
msp_micro_working[, age_factor := relevel(age_factor, '25')]
mod <- lm(hhincome~age_factor, data=msp_micro_working)
out <- as.data.table(avg_predictions(mod, by='age_factor'))
out[, age := as.numeric(as.character(age_factor))]
ggplot(data=out, aes(x=age,y=estimate,ymin=`conf.low`,ymax=`conf.high`)) + 
  geom_linerange() + 
  geom_point(size=2) + 
  labs(x='Age', y='Household income') + 
  theme_bw()
```

---

\frametitle{Statistical inference}

\tiny
```{r, echo=F}
msp_micro <- fread('https://raw.githubusercontent.com/ngraetz/soc5811/refs/heads/main/data/acs2023_hennepin.csv')
ggplot(data=msp_micro[educ=='4 years of college' & age %in% 25:64 & !is.na(hhincome),],
       aes(x=age,y=hhincome)) + 
  geom_hex(alpha=0.7) + 
  geom_smooth(method='lm',color='black') +
  scale_fill_viridis_c(option='plasma') + 
  labs(x='Age', y='Household income') + 
  theme_bw()
```

---

\frametitle{Statistical inference}

>- Could the sample relationship be due to chance?

>- Based on our _sample_, how sure are we that the _population_ parameter is different from 0?

---

\frametitle{Statistical inference}

>- \textbf{Parameters} = the unknown numbers that determine a statistical model. 

>- Parameters can be used to simulate new data from the model.

---

\frametitle{Statistical inference}

>- \textbf{Statistical inference} = a set of operations on data that yield estimates and uncertainty statements about \textbf{predictions and parameters} of some underlying process or population.

>- More simply, we observe data and we try to learn something about where it came from (i.e., the data generating process).

>- \textbf{This involves uncertainty.} 

---

\frametitle{Statistical inference}

* Say I have a population of 1000 people and I ask 30 of them their income. The parameter I'm interested in inferring something about is the average income in the entire population.

---

\frametitle{Statistical inference}

\tiny
```{r, echo=FALSE}
test <- data.table(income=round(rnorm(30, 65, 20)))
ggplot(data=test,
       aes(x=income)) + 
  geom_histogram(color=NA,fill='grey') + 
  geom_vline(xintercept = test[,mean(income)], color='red', size=2) + 
  labs(x='Income',y='Count') + 
  theme_bw()
```

---

\frametitle{Statistical inference}

\tiny
```{r, echo=FALSE}
get_sample <- function(i) {
  test <- data.table(income=round(rnorm(30, 65, 20)))
  test[, mean := mean(income)]
  test[, sample := paste0('Sample ',i)]
  return(test)
}
test <- rbindlist(lapply(1:6, get_sample))
ggplot(data=test,
         aes(x=income)) + 
    geom_histogram(color=NA,fill='grey') + 
    geom_vline(aes(xintercept=mean), color='red', size=1) + 
    facet_wrap(~sample, nrow=2) + 
    labs(x='Income',y='Count') + 
    theme_bw()
```

---

\frametitle{Sampling distribution}

>- \textbf{Sampling distribution} = the set of possible datasets that could have been observed if the data collection process had been re-done, along with the probabilities of these values.

>- In practice, we will not know the sampling distribution; we can only estimate it. 

---

\frametitle{Standard errors}

>- \textbf{Standard deviation} = a measure of the spread of data around the mean. 

$$\sigma=\sqrt{\frac{\sum(x_i-\mu)^2}{N}}$$

\begin{center}
\includegraphics[width = 0.7\textwidth]{images/normal_sd.png}
\end{center}

---

\frametitle{Standard errors}

* \textbf{Standard deviation} = a measure of the spread of data around the mean. 

$$\sigma=\sqrt{\frac{\sum(x_i-\mu)^2}{N}}$$

* \textbf{Standard error} = ???

---

\frametitle{Standard errors}

* \textbf{Standard deviation} = a measure of the spread of data around the sample mean. 

$$\sigma=\sqrt{\frac{\sum(x_i-\mu)^2}{N}}$$

* \textbf{Standard error} = a measure of the spread of sample means around the population mean; the standard deviation of the sampling distribution. 

---

\frametitle{Sampling distribution}

\tiny
```{r, echo=FALSE}
get_sample <- function(i) {
  test <- data.table(income=round(rnorm(30, 65, 20)))
  gg <- ggplot(data=test,
         aes(x=income)) + 
    geom_histogram(color=NA,fill='grey') + 
    geom_vline(xintercept = test[,mean(income)], color='red', size=1) + 
    labs(x='Income',y='Count') + 
    theme_bw()
  return(gg)
}
test <- lapply(1:6, get_sample)
plot_grid(plotlist = test, nrow=2)
```

---

\frametitle{Sampling distribution}

Sampling distribution of the mean:

\tiny
```{r, echo=FALSE}
get_sample <- function(i) {
  test <- data.table(income=round(rnorm(30, 65, 20)))
  return(mean(test[,income]))
}
test <- unlist(lapply(1:1000, get_sample))
test <- data.table(income=test)
ggplot(data=test,
         aes(x=income)) + 
    geom_histogram(color=NA,fill='grey') + 
    geom_vline(xintercept = test[,mean(income)], color='red', size=1) + 
    labs(x='Income',y='Count') + 
    theme_bw()
```

---

\frametitle{Sampling distribution}

What do you notice about the shape of this distribution?

\tiny
```{r, echo=FALSE}
ggplot(data=test,
         aes(x=income)) + 
    geom_histogram(color=NA,fill='grey') + 
    geom_vline(xintercept = test[,mean(income)], color='red', size=1) + 
    labs(x='Income',y='Count') + 
    theme_bw()
```

---

\frametitle{Sampling distribution}

* \textbf{Central Limit Theorem}

\tiny
```{r, echo=FALSE}
ggplot(data=test,
         aes(x=income)) + 
    geom_histogram(color=NA,fill='grey') + 
    geom_vline(xintercept = test[,mean(income)], color='red', size=1) + 
    labs(x='Income',y='Count') + 
    theme_bw()
```

---

\frametitle{Sampling distribution}

\tiny
```{r, echo=FALSE}
get_sample_size <- function(size) {
  get_sample <- function(i,size) {
    test <- data.table(income=round(rnorm(size, 65, 20)))
    return(mean(test[,income]))
  }
  test <- unlist(lapply(1:1000, get_sample, size=size))
  test <- data.table(income=test)
  test[, size := size]
  return(test)
}
test2 <- rbindlist(lapply(c(30,100,400), get_sample_size))
ggplot(data=test2,
         aes(x=income)) + 
    geom_histogram(color=NA,fill='grey') + 
    facet_wrap(~size,nrow=1) + 
    labs(x='Income',y='Count') + 
    theme_bw()
```

---

\frametitle{Sampling distribution}

How do I interpret the standard deviation of this distribution?

\tiny
```{r, echo=FALSE}
ggplot(data=test,
         aes(x=income)) + 
    geom_histogram(color=NA,fill='grey') + 
    geom_vline(xintercept = test[,mean(income)], color='red', size=1) + 
    labs(x='Income',y='Count') + 
    theme_bw()
```

---

\frametitle{Sampling distribution}

>- However defined, the standard error is a measure of the variation in an estimate and gets smaller as sample size gets larger, converging on zero as the sample increases in size.

>- The \textbf{confidence interval} represents a range of values of a parameter or quantity of interest that are roughly consistent with the data, given the assumed sampling distribution.

---

\frametitle{Sampling distribution}

\tiny
```{r, echo=FALSE}
ggplot(data=test,
         aes(x=income)) + 
    geom_histogram(color=NA,fill='grey') + 
    geom_vline(xintercept = test[,mean(income)], color='red', size=1) + 
    geom_vline(xintercept = test[,mean(income)+1.96*sd(income)], color='red', size=1, linetype='dashed') +
    geom_vline(xintercept = test[,mean(income)-1.96*sd(income)], color='red', size=1, linetype='dashed') +
    labs(x='Income',y='Count') + 
    theme_bw()
```

---

\frametitle{Bias}

It is possible to be very confidently wrong!

```{r, echo=FALSE}
test <- data.table(income=round(rnorm(10000, 65, 20)))
ggplot(data=test,
         aes(x=income)) + 
    geom_histogram(color=NA,fill='grey') + 
    labs(x='Income',y='Count') + 
    theme_bw()
```

---

\frametitle{Bias}

Say only people with incomes above 50k complete my survey:

```{r, echo=FALSE}
ggplot(data=test,
         aes(x=income)) + 
    geom_histogram(color=NA,fill='grey') + 
    geom_histogram(data=test[income>50,], color=NA,fill='red') + 
    labs(x='Income',y='Count') + 
    theme_bw()
```

---

\frametitle{Bias}

\textbf{Describing uncertainty:}

>- Standard errors capture a particular (but very important!) type of uncertainty: sampling error.

>- There are lots of other reasons to be uncertain! 

---

\frametitle{Statistical significance}

What can we do with standard errors?

* Hypothesis testing


