---
title: "SOC-5811 Week 6: Simulation"
author: "Nick Graetz"
date: "10/6/2025"
output: 
    beamer_presentation:
        incremental: false
        includes: 
            in_header: Minnesota.tex
---

```{r setup, include=FALSE}
## Libraries
library(tidyverse)
library(data.table)
library(ggplot2)
library(haven)
library(formatR)
library(marginaleffects)
library(cowplot)
# knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 20), tidy = TRUE)
knitr::opts_chunk$set(
 fig.width = 6,
 fig.height = 4,
 warning = FALSE, 
 message = FALSE
)
dropbox <- 'C:/Users/ncgra/Dropbox/'
fig_dir <- paste0(dropbox,
                  'Minnesota/repos/soc5811/figures/')
data_dir <- paste0(dropbox,
                   'Minnesota/repos/soc5811/data/')
```

---

\frametitle{Review}

>- Sampling distributions

>- Statistical inference and p-values

---

\frametitle{Review}

* \textbf{Standard deviation} = a measure of the spread of data around the sample mean. 

$$\sigma=\sqrt{\frac{\sum(x_i-\mu)^2}{n}}$$

* \textbf{Standard error} = a measure of the spread of sample means around the population mean; the standard deviation of the sampling distribution. 

$$\text{Standard error}=\frac{\sigma}{\sqrt{n}}$$

---

\frametitle{Simulation}

>- Last week was \textbf{statistical inference}: We _observe_ sample data $\longrightarrow$ we estimate the most likely _unknown_ population model that produced that sample data.

>- This week is \textbf{simulation}: We _know_ the population model $\longrightarrow$ we simulate all the sample data is could plausibly produce (i.e., we simulate the _sampling distribution_). 

---

\frametitle{Simulation}

Why is simulation an important concept/skill? 

>- Develop statistical intuition without needing math.

>- We can use simulation to approximate the sampling distribution of data and propagate this to the sampling distribution of statistical estimates (aka \textbf{bootstrapping}).

>- Very important principle for more complex models: you can always simulate.

---

\frametitle{Simulation}

>- Remember: this is all about understanding how we do \textbf{statistical inference}.

>- What can I learn about the population model from the data I observe? 

>- Are my sample parameters real, or could they plausibly have been observed by chance?

---

\frametitle{Simulation}

\begin{center}
\includegraphics[width = 0.5\textwidth]{images/table2_example.png}
\end{center}

---

\frametitle{Simulation}

\begin{center}
\includegraphics[width = 0.5\textwidth]{images/table2_example2.png}
\end{center}

---

\frametitle{This week: R skills}

>- Functions in R

>- Simulating in R

>- Bootstrapping in R

>- Introduction to multivariable models

---

\frametitle{Functions in R}

\small
```{r, echo=T}
log(5)
```

---

\frametitle{Functions in R}

\small
```{r, echo=T}
new_function <- function(x) {
  y <- x*10
  return(y)
}
```

---

\frametitle{Functions in R}

\small
```{r, echo=T}
new_function <- function(x) {
  y <- x*10
  return(y)
}
new_function(5)
```

---

\frametitle{Functions in R}

\small
```{r, echo=T}
new_function <- function(x) {
  y <- x*10
  return(y)
}
sapply(c(5,10,15,20), new_function)
```

---

\frametitle{Simulation}

>- Random variables

>- Distributions:
>   - Normal distribution
>   - Binomial distribution

>- \textbf{Simulation of random variables in R.}

---

\frametitle{Simulation}

X = the number of heads I get if I flip a coin 100 times.

X = Binomial(n=100, p=0.5)

\small
```{r, echo=TRUE}
rbinom(n=1, size=100, prob=0.5)
```

---

\frametitle{Simulation}

\small
```{r, echo=TRUE}
rbinom(n=1, size=100, prob=0.5)
rbinom(n=1, size=100, prob=0.5)
rbinom(n=1, size=100, prob=0.5)
```

---

\frametitle{Simulation}

\small
```{r, echo=TRUE}
trials <- rbinom(n=10000, size=100, prob=0.5)
trials <- data.table(heads=trials)
head(trials)
```

---

\frametitle{Simulation}

\small
```{r, echo=F}
ggplot(data=trials,
       aes(x=heads)) + 
  geom_density(fill='lightgrey') + 
  theme_bw()
```

---

\frametitle{Simulation}

\small
```{r, echo=F}
ggplot(data=trials,
       aes(x=heads)) + 
  geom_density(fill='lightgrey') + 
  geom_vline(xintercept = 50, color='red', size=1) + 
  theme_bw()
```


---

\frametitle{Simulation}

Imagine flipping a coin 100 times and interpreting the results relative to some null probability model.

Null model: $P(\text{heads})=0.5$; this is a fair coin.

\textbf{Results: 55 heads, 45 tails.}

---

\frametitle{Simulation}

```{r, echo=F}
ggplot(data=trials,
       aes(x=heads)) + 
  geom_density(fill='lightgrey') +
  geom_vline(xintercept = 55, color='red', size=1) + 
  theme_bw()
```

---

\frametitle{Simulation}

```{r, echo=T}
pbinom(55, 100, 0.5, lower.tail = FALSE)
```

---

\frametitle{Simulation}

Imagine flipping a coin 100 times and interpreting the results relative to some null probability model.

Null model: $P(\text{heads})=0.5$; this is a fair coin.

\textbf{Results: 55 heads, 45 tails.}

\textbf{Is the hypothesis that this is a fair coin consistent with our data?}
\textbf{Could a fair coin plausibly produce our data?}

---

\frametitle{Simulation}

```{r, echo=F}
ggplot(data=trials,
       aes(x=heads)) + 
  geom_density(fill='lightgrey') +
  geom_vline(xintercept = 70, color='red', size=1) + 
  theme_bw()
```

---

\frametitle{Simulation}

```{r, echo=T}
pbinom(70, 100, 0.5, lower.tail = FALSE)
```

---

\frametitle{Simulation}

\small
```{r, echo=TRUE}
mean_of_normal_trials <- function(i, ss) {
  mean <- mean(rnorm(ss, 0, 5))
  return(mean)
}
normal_trials <- function(ss) {
  trials <- lapply(1:1000, 
                   mean_of_normal_trials, 
                   ss=ss)
  trials <- data.table(sample_size=ss,
                       mean=unlist(trials))
  return(trials)
}
trials <- lapply(c(30,100,400), normal_trials)
trials <- rbindlist(trials)
```

---

\frametitle{Simulation}

\tiny
```{r, echo=F}
ggplot(data=trials,
         aes(x=mean)) + 
    geom_histogram(color=NA,fill='grey') + 
    facet_wrap(~sample_size,nrow=1) + 
    labs(x='Difference',y='Count') + 
    theme_bw()
```

---

\frametitle{Simulation}

\tiny
```{r, echo=F}
ggplot(data=trials,
         aes(x=mean)) + 
    geom_histogram(color=NA,fill='grey') + 
    geom_vline(xintercept = 1, size=1, color='red') + 
    facet_wrap(~sample_size,nrow=1) + 
    labs(x='Difference',y='Count') + 
    theme_bw()
```

---

\frametitle{Bootstrapping}

* Simulating a sampling distribution when you \textbf{do not know} the underlying probability model.

* Randomly sample the \textbf{observed} data and calculate your statistic many times ("bootstrapping").

---

\frametitle{Bootstrap}

\tiny
```{r, echo=TRUE}
psid <- fread('https://raw.githubusercontent.com/ngraetz/soc5811/refs/heads/main/data/psid.csv')
psid <- psid %>%
    filter(g3_house_value!=0)
```

---

\frametitle{Bootstrap}

\small
```{r, echo=TRUE}
mean_analytic <- psid %>% 
  summarise(mean=mean(g3_house_value)) %>% 
  pull(mean)
se_analytic <- psid %>% 
  summarise(se=sd(g3_house_value)/sqrt(nrow(psid))) %>%
  pull(se)
mean_analytic
se_analytic
```

---

\frametitle{Bootstrap}

\small
```{r, echo=T}
get_sample <- function(i) {
  random_sample <- psid %>%
    sample_n(size=nrow(psid),replace=T) %>%
    summarise(mean=mean(g3_house_value))
  return(random_sample)
}
samples <- rbindlist(lapply(1:100,get_sample))
```

---

\frametitle{Bootstrap}

```{r, echo=F}
ggplot(data=samples,
       aes(x=mean)) + 
  geom_density(fill='lightgrey') + 
  theme_bw()
```

---

\frametitle{Bootstrap}

\small
```{r, echo=TRUE}
## Calculate standard error
se_bootstrap <- samples %>%
  summarise(sd=sd(mean)) %>%
  pull(sd)

## Compare
se_analytic
se_bootstrap
```

---

\frametitle{Bootstrap}

\small
```{r, echo=FALSE}
out <- data.table(model=c('Analytic','Bootstrap'),
                  estimate=c(mean_analytic,mean_analytic),
                  lower=c(mean_analytic-(2*se_analytic),mean_analytic-(2*se_bootstrap)),
                  upper=c(mean_analytic+(2*se_analytic),mean_analytic+(2*se_bootstrap))) 
ggplot(data=out,
       aes(y=model,x=estimate,xmin=lower,xmax=upper)) +
  geom_linerange() + 
  geom_point(size=3) + 
  labs(y='',x='Estimate') + 
  theme_bw() 
```

---

\frametitle{Bootstrap}

\small
```{r, echo=F}
psid_subset <- psid %>%
  filter(!is.na(g1_edu_cat) & g3_house_value!=0) %>% 
  mutate(above_hs=ifelse(g1_edu_cat %in% c('less_hs','hs'), 'less', 'more'))
```


\small
```{r, echo=TRUE}
psid_subset <- psid %>%
  filter(!is.na(g1_edu_cat) & g3_house_value!=0) %>% 
  mutate(above_hs=ifelse(g1_edu_cat %in% c('less_hs','hs'), 
                         'less', 'more'))

get_sample <- function(i) {
  random_sample <- psid_subset %>%
    sample_n(size=nrow(psid_subset),replace=T) 
  model <- lm(g3_house_value_percentile~above_hs,
            data=random_sample)
  out <- avg_slopes(model)
  return(out)
}
coef_samples <- rbindlist(lapply(1:100,get_sample))
```

---

\frametitle{Bootstrap}

```{r, echo=F}
ggplot(data=coef_samples,
       aes(x=estimate)) + 
  geom_density(fill='lightgrey') + 
  theme_bw()
```

---

\frametitle{Bootstrap}

\small
```{r, echo=TRUE}
## Compare to lm()
model <- lm(g3_house_value_percentile~above_hs,
            data=psid_subset)
analytic_coefs <- as.data.table(avg_slopes(model))
```

---

\frametitle{Bootstrap}

\tiny
```{r, echo=F}
out <- data.table(model=c('Analytic','Bootstrap'),
                  estimate=c(analytic_coefs$estimate,analytic_coefs$estimate),
                  lower=c(analytic_coefs$conf.low,analytic_coefs$estimate-(2*sd(coef_samples$estimate))),
                  upper=c(analytic_coefs$conf.high,analytic_coefs$estimate+(2*sd(coef_samples$estimate)))) 
ggplot(data=out,
       aes(y=model,x=estimate,xmin=lower,xmax=upper)) +
  geom_vline(xintercept = 0, linetype='dashed') + 
  geom_linerange() + 
  geom_point(size=3) + 
  labs(y='',x='Estimate') + 
  theme_bw()
```

