---
title: "SOC-5811 Week 7: Regression with multiple variables"
author: "Nick Graetz"
date: "10/13/2025"
output: 
    beamer_presentation:
        incremental: false
        includes: 
            in_header: Minnesota.tex
---

```{r setup, include=FALSE}
## Libraries
library(tidyverse)
library(data.table)
library(ggplot2)
library(haven)
library(formatR)
library(marginaleffects)
library(cowplot)
library(fastDummies)
# knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 20), tidy = TRUE)
knitr::opts_chunk$set(
 fig.width = 6,
 fig.height = 4,
 warning = FALSE, 
 message = FALSE
)
dropbox <- 'C:/Users/ngraetz/Dropbox/'
fig_dir <- paste0(dropbox,
                  'Minnesota/repos/soc5811/figures/')
data_dir <- paste0(dropbox,
                   'Minnesota/repos/soc5811/data/')
```

---

\frametitle{Review categorical predictor variables}

\tiny
```{r, echo=T}
penguin_df <- 
  palmerpenguins::penguins %>%
  na.omit()
head(penguin_df)
```

---

\frametitle{Review categorical predictor variables}

What happens if we use a categorical predictor variable with \textbf{two} levels? 

$$y_i=\beta_0+\beta_1(\text{Sex})$$

---

\frametitle{Review categorical predictor variables}

\tiny
```{r, echo=T}
summary(lm(bill_depth_mm ~ sex, data=penguin_df))
```

---

\frametitle{Review categorical predictor variables}

\tiny
```{r, echo=T}
penguin_df <- penguin_df %>%
 dummy_cols('sex')
penguin_df %>% 
  select(species,bill_length_mm,bill_depth_mm,sex,sex_female,sex_male)
```

---

\frametitle{Review categorical predictor variables}

\tiny
```{r, echo=T}
summary(lm(bill_depth_mm ~ sex_male, data=penguin_df))
```

---

\frametitle{Review categorical predictor variables}

>- Do I need to add both dummy variables to my regression model? Why or why not?

\tiny
```{r, echo=T}
penguin_df %>% 
  select(species,bill_length_mm,bill_depth_mm,sex,sex_female,sex_male)
```

---

\frametitle{Review categorical predictor variables}

\tiny
```{r, echo=T}
summary(lm(bill_depth_mm ~ sex_male + sex_female, data=penguin_df))
```

---

\frametitle{Review multicollinearity}

>- The problem of \textbf{multicollinearity}.

>- In a regression model, you can't have predictors that are a perfect function of one another. In other words, you can't have perfect \textbf{multicollinearity} where predictors are perfectly correlated with one another. 

>- Why?

---

\frametitle{Review}

1. Choose a functional form for the model.

$$\text{depth}_i=\beta_0+\beta_1{\text{sex\_male}}+\beta_2{\text{sex\_female}}+\epsilon_i$$

2. Choose an estimator (i.e., objective function) that relates the model to real observed data (e.g., OLS).

$$\text{Sum of squared residuals}=\sum_{i=1}^n (y_i-\widehat{\beta_0}-\widehat{\beta_1}x_i)^2$$

3. Fit the model (e.g., estimate parameters) by optimizing the objective function.

---

\frametitle{Review multicollinearity}

\begin{center}
\includegraphics[width = 0.8\textwidth]{images/ss_example.png}
\end{center}

---

\frametitle{Review multicollinearity}

>- If you provide a categorical predictor variable to lm(), the default behavior is to transform it into N-1 dummy variables where N is the number of unique levels.

---

\frametitle{Review categorical predictor variables}

\tiny
```{r, echo=T}
penguin_df <- penguin_df %>%
 dummy_cols('species')
head(penguin_df)
```

---

\frametitle{Review categorical predictor variables}

\tiny
```{r, echo=T}
summary(lm(bill_depth_mm ~ species_Adelie + species_Gentoo + species_Chinstrap, 
           data=penguin_df))
```

---

\frametitle{Review categorical predictor variables}

\tiny
```{r, echo=T}
penguin_df <- penguin_df %>%
  mutate(species = factor(species, levels=c('Chinstrap','Adelie','Gentoo')))
summary(lm(bill_depth_mm ~ species, data = penguin_df))
```

---

\frametitle{Review categorical predictor variables}

$$y_i=\beta_0+\beta_1(\text{Adelie}) + \beta_2(\text{Gentoo})$$

---

\frametitle{Review categorical predictor variables}

$$y_i=\beta_0+\beta_1(\text{Adelie}) + \beta_2(\text{Gentoo})$$

$$y_i=(18.42)+(-0.07)(\text{Adelie}) + (-3.42)(\text{Gentoo})$$

---

\frametitle{Review categorical predictor variables}

$$y_i=\beta_0+\beta_1(\text{Adelie}) + \beta_2(\text{Gentoo})$$

$$y_i=(18.42)+(-0.07)(\text{Adelie}) + (-3.42)(\text{Gentoo})$$

$$E[Y|\text{species}=\text{Adelie}] = (18.42)+(-0.07)(1) + (-3.42)(0)=\bm{18.35}$$

---

\frametitle{Review categorical predictor variables}

$$y_i=\beta_0+\beta_1(\text{Adelie}) + \beta_2(\text{Gentoo})$$

$$y_i=(18.42)+(-0.07)(\text{Adelie}) + (-3.42)(\text{Gentoo})$$

$$E[Y|\text{species}=\text{Adelie}] = (18.42)+(-0.07)(1) + (-3.42)(0)=\bm{18.35}$$

$$E[Y|\text{species}=\text{Chinstrap}] = (18.42)+(-0.07)(0) + (-3.42)(0)=\bm{18.42}$$

---

\frametitle{Multiple predictors}

>- The coefficient $\beta_k$ is the average or expected difference in outcome y, comparing two people who
differ by one unit in the predictor $x_k$ while being equal in all the other predictors. This is sometimes
stated in shorthand as comparing two people (or, more generally, two observational units) that differ
in $x_k$ with all the other predictors held constant.

---

\frametitle{Multiple predictors}

\tiny
```{r, echo=T}
mod <- lm(bill_depth_mm ~ bill_length_mm + species, data = penguin_df)
summary(mod)
```

---

\frametitle{Multiple predictors}

>- $\beta_0=8.63$: If a penguin is a Chinstrap penguin with a bill length of 0, then we would predict this penguin to have a bill depth of \textbf{8.63.}
>- $\beta_1=0.20$: Comparing penguins within the same species but whose bill lengths differ by 1, we would predict their difference in bill depths to be \textbf{0.20.}
>- $\beta_2=1.93$: Comparing penguins who have the same bill length but one is Chinstrap and one is Adelie, we would predict their difference in bill depths to be \textbf{1.93.} 
>- $\beta_3=-3.17$: Comparing penguins who have the same bill length but one is Chinstrap and one is Gentoo, we would predict their difference in bill depths to be \textbf{3.17.} 

---

\frametitle{Review}

>- On the board: health as a function of income and wealth. 

>- The coefficients of each variable are capturing variation _independent_ of the other variable, or _marginal_ to the other variable. To estimate the coefficient of income on health, we are using variation in our data that is mechanically unrelated to wealth; in other words, differences \textbf{cannot} be attributable to wealth because we are only comparing units with the same level of wealth.

---

\frametitle{Review}

>- $\text{health}=\beta_0+\beta_1\text{income}+\beta_2\text{wealth}$

>- The effects of each variable are also _independent_ of each other. My estimated coefficient of income is the same no matter what level of wealth I'm using. I'm averaging over the differences within any given level of wealth to arrive at an overall _average_ coefficient for income on health. 

>- We may be averaging over interesting variation. We could imagine the effect of income on health _depends_ on level of wealth. We may want to estimate an _interaction_ between income and wealth. 

---

\frametitle{Multiple predictors}

>- Regression is a way of making comparisons; it is very important to understand and be able to articulate the comparison you are making.

$$d_i=\beta_0+\beta_1(l_i)+\beta_2(s^\text{Adelie}_i)+\beta_3(s^\text{Gentoo}_i)$$

---

\frametitle{Multiple predictors}

>- Regression is a way of making comparisons; it is very important to understand and be able to articulate the comparison you are making.

$$d_i=\beta_0+\beta_1(l_i)+\beta_2(s^\text{Adelie}_i)+\beta_3(s^\text{Gentoo}_i)$$

$$\beta_1=E[d|l=x+1,s]-E[d|l=x,s]=\bm{0.20}$$

---

\frametitle{Multiple predictors}

>- Regression is a way of making comparisons; it is very important to understand and be able to articulate the comparison you are making.

$$d_i=\beta_0+\beta_1(l_i)+\beta_2(s^\text{Adelie}_i)+\beta_3(s^\text{Gentoo}_i)$$

$$\beta_1=E[d|l=x+1,s]-E[d|l=x,s]=\bm{0.20}$$

$$\beta_2=E[d|s=\text{Adelie}, l]-E[d|s=\text{Chinstrap}, l]=\bm{1.93}$$

---

\frametitle{Multiple predictors}

>- Regression is a way of making comparisons; it is very important to understand and be able to articulate the comparison you are making.

$$d_i=\beta_0+\beta_1(l_i)+\beta_2(s^\text{Adelie}_i)+\beta_3(s^\text{Gentoo}_i)$$

$$\beta_1=E[d|l=x+1,s]-E[d|l=x,s]=\bm{0.20}$$

$$\beta_2=E[d|s=\text{Adelie}, l]-E[d|s=\text{Chinstrap}, l]=\bm{1.93}$$

$$\beta_3=E[d|s=\text{Gentoo}, l]-E[d|s=\text{Chinstrap}, l]=\bm{-3.17}$$

---

$$d_i=\beta_0+\beta_1(l_i)+\beta_2(s^\text{Adelie}_i)+\beta_3(s^\text{Gentoo}_i)$$

>- How many estimates do we have of the association between bill length and depth? 

>- We just have one estimate ($\beta_1$) regardless of whether we control for species.

---

\frametitle{Holding something constant}

\tiny
$$d_i=\beta_0+\beta_1(l_i)$$

\tiny
```{r, echo=F}
penguin_df <- penguin_df %>% 
  group_by(species) %>% 
  mutate(bill_depth_mm_demeaned = 
           bill_depth_mm - mean(bill_depth_mm),
         bill_length_mm_demeaned = 
           bill_length_mm - mean(bill_length_mm))
ggplot(data=penguin_df,
       aes(x=bill_length_mm, y=bill_depth_mm, fill=species)) + 
  geom_point(shape=21,size=3) + 
  geom_smooth(aes(x=bill_length_mm, y=bill_depth_mm), method='lm',se=F,size=1, color='red', inherit.aes = F) +
  scale_fill_viridis_d() + 
  labs(x="Length", y="Depth", fill='Species') +
  theme_bw()
```

---

\frametitle{Holding something constant}

\tiny
$$d_i=\beta_0+\beta_1(l_i)+\beta_2(s^\text{Adelie}_i)+\beta_3(s^\text{Gentoo}_i)$$

\tiny
```{r, echo=F}
ggplot(data=penguin_df,
       aes(x=bill_length_mm_demeaned, y=bill_depth_mm_demeaned, fill=species)) + 
  geom_point(shape=21,size=3) + 
  geom_smooth(aes(x=bill_length_mm_demeaned, y=bill_depth_mm_demeaned), method='lm',se=F,size=1, color='red', inherit.aes = F) +
  scale_fill_viridis_d() + 
  labs(x="Length", y="Depth", fill='Species') +
  theme_bw()
```

---

\frametitle{Multiple predictors}

\tiny
$$d_i=\beta_0+\beta_1(l_i)+\beta_2(s^\text{Adelie}_i)+\beta_3(s^\text{Gentoo}_i)$$

\tiny
```{r, echo=F}
mod <- lm(bill_depth_mm ~ bill_length_mm + species, data = penguin_df)
predictions <- tibble(depth_preds = predictions(mod)$estimate)
penguin_df <- penguin_df %>%
  bind_cols(., predictions)
ggplot(data=penguin_df,
       aes(x=bill_length_mm, y=bill_depth_mm, fill=species)) + 
  geom_point(shape=21,size=3) + 
  geom_line(aes(y=depth_preds),size=1,color='red') + 
  scale_fill_viridis_d() + 
  labs(x="Length", y="Depth", fill='Species') +
  theme_bw()
```

---

\frametitle{Interactions}

>- In our model, the slope of the regression for bill length was forced to be equal across all subgroups defined by species. In other words, we just have one $\beta_1$. 

$$d_i=l_i+s_i$$

$$d_i=\beta_0+\beta_1(l_i)+\beta_2(s^\text{Adelie}_i)+\beta_3(s^\text{Gentoo}_i)$$

---

\frametitle{Interactions}

>- Now how many coefficients do we have for length?

$$d_i=l_i+s_i+(l_i\times s_i)$$

$$d_i=\beta_0+\beta_1(l_i)+\beta_2(s^\text{Adelie}_i)+\beta_3(s^\text{Gentoo}_i)+\beta_4(l_i \times s^\text{Adelie}_i)+\beta_5(l_i \times s^\text{Gentoo}_i)$$

---

\frametitle{Interactions}

\tiny
```{r, echo=T}
mod2 <- lm(bill_depth_mm ~ bill_length_mm * species, data = penguin_df)
summary(mod2)
```

---

\frametitle{Interactions}

\tiny
```{r, echo=F}
predictions2 <- tibble(depth_preds2 = predictions(mod2)$estimate)
penguin_df <- penguin_df %>%
  bind_cols(., predictions2)
ggplot(data=penguin_df,
       aes(x=bill_length_mm, y=bill_depth_mm, fill=species)) + 
  geom_point(shape=21,size=3) + 
  geom_line(aes(y=depth_preds),size=1,color='red') + 
  geom_line(aes(y=depth_preds2),size=1,color='black') + 
  scale_fill_viridis_d() + 
  labs(x="Length", y="Depth", fill='Species') +
  theme_bw()
```

---

\frametitle{Interactions}

\tiny
```{r, echo=T}
mod3 <- lm(bill_depth_mm ~ bill_length_mm * body_mass_g, data = penguin_df)
summary(mod3)
```

---

\frametitle{Review: Why multiple variables?}

>- Social process are typically not the function of one independent variable

>- Simpsons paradox

>- "Holding constant" = "Comparing two items $i$ and $j$ that differ by an amount $x$ on predictor $k$ but are identical on all other predictors, the predicted difference $y_i - y_j$ is $B_k(x)$, on average."

---

\frametitle{Review: Why multiple variables?}

Different research questions:

>- What is the average association between grandparent wealth and grandchild wealth?
>- What is the difference in grandchild wealth by grandparent race?
>- What is the average association between grandparent wealth and grandchild wealth conditional on race (i.e., comparing grandparents within the same race)? 
>- What is the difference in the average association between grandparent wealth and grandchild wealth by race?

---

\frametitle{Confounding}

>- Review Simpson's Paradox

\begin{center}
\includegraphics[width = 0.8\textwidth]{images/simpsons2.png}
\end{center}

---

\frametitle{Confounding}

$$\text{Admission}=\beta_0+\beta_1(\text{Gender})$$

$$\text{Admission}=\beta_0+\beta_1(\text{Gender})+\beta_2(\text{Department})$$

---

\frametitle{Confounding}

>- Do not listen to anyone who says "Well, you _have_ to control for X." 

>- What they are really saying: "The comparison in which you are not controlling for X is not useful; I'm assuming you want to make the comparison where you control for X." This can be an argument on theoretical grounds about the utility of various research questions.

>- There may be very good reasons to control for X! But it _completely_ depends on your research question; more specifically, exactly what \textbf{comparison} you say you're trying to make. 

>- We are transitioning to causal inference and the concept of _confounding_. 

---

\frametitle{Confounding}

>- I'm curious whether people who drink more wine live longer.

>- I do a huge survey where I follow people over time and track how much wine they drink and if they die.

>- I fit the regression model: $\text{mortality rate}=\beta_0+\beta_1\text{wine}$.

>- People who drink more wine are less likely to die (p<0.05)!

>- People who drank more wine are less likely to die _because_ they drank more wine; _if_ people drink more wine, _then_ they will be less likely to die. 

---

\frametitle{Confounding}

\small
$$\text{mortality rate}=\beta_0+\beta_1\text{wine}+\beta_2\text{age}+\beta_3\text{income}+\beta_4\text{education}+\beta_5\text{social}$$

>- People who drink more wine are less likely to die comparing within levels of age, income, education, and social networks (p<0.05)!

---

\frametitle{Confounding}

$$\text{mortality rate}=\beta_0+\beta_1\text{wine}$$

How might we avoid having to control for so many things? In other words, how might we guarantee the _only_ difference across our comparisons of drinking is the drinking?

