---
title: "SOC-5811 Week 9: Potential outcomes"
author: "Nick Graetz"
date: "10/27/2025"
output: 
    beamer_presentation:
        incremental: false
        includes: 
            in_header: Minnesota.tex
---

```{r setup, include=FALSE}
library(data.table)
library(knitr)
library(kableExtra)
library(arrow)
library(haven)
library(ggplot2)
library(marginaleffects)
library(tidyverse)
knitr::opts_chunk$set(echo = FALSE)
dropbox <- 'C:/Users/ncgra/Dropbox/'
fig_dir <- paste0(dropbox, 'Minnesota/repos/soc5811/figures/')
```

---

\frametitle{Review}

\textbf{Research Question:} What is the difference in grandchild wealth by grandparent race, controlling for grandparent wealth?

* $Y$ = grandchild wealth (continuous)
* $X$ = grandparent wealth (continuous)
* $Z$ = grandparent race (categorical: Black, White)

\begin{align}
  &\textbf{A: } y_i=\beta_0+\boxed{\beta_1}(X) \\
  &\textbf{B: } y_i=\beta_0+\boxed{\beta_1}(X)+\beta_2(Z) \\
  &\textbf{C: } y_i=\beta_0+\beta_1(X)+\boxed{\beta_2}(Z) \\
  &\textbf{D: } y_i=\beta_0+\beta_1(X)+\beta_2(Z)+\boxed{\beta_3}(X \times Z) 
\end{align}

---

\frametitle{Review}

\textbf{Research Question:} What is the difference in the average association between grandparent wealth and grandchild wealth by race?

* $Y$ = grandchild wealth (continuous)
* $X$ = grandparent wealth (continuous)
* $Z$ = grandparent race (categorical: Black, White)

\begin{align}
  &\textbf{A: } y_i=\beta_0+\boxed{\beta_1}(X) \\
  &\textbf{B: } y_i=\beta_0+\boxed{\beta_1}(X)+\beta_2(Z) \\
  &\textbf{C: } y_i=\beta_0+\beta_1(X)+\boxed{\beta_2}(Z) \\
  &\textbf{D: } y_i=\beta_0+\beta_1(X)+\beta_2(Z)+\boxed{\beta_3}(X \times Z) 
\end{align}

---

\frametitle{Review}

\textbf{Research Question:} What is the average association between grandparent wealth and grandchild wealth?

* $Y$ = grandchild wealth (continuous)
* $X$ = grandparent wealth (continuous)
* $Z$ = grandparent race (categorical: Black, White)

\begin{align}
  &\textbf{A: } y_i=\beta_0+\boxed{\beta_1}(X) \\
  &\textbf{B: } y_i=\beta_0+\boxed{\beta_1}(X)+\beta_2(Z) \\
  &\textbf{C: } y_i=\beta_0+\beta_1(X)+\boxed{\beta_2}(Z) \\
  &\textbf{D: } y_i=\beta_0+\beta_1(X)+\beta_2(Z)+\boxed{\beta_3}(X \times Z) 
\end{align}

---

\frametitle{Review DAGs}

\begin{figure}[h]
\centering

	\large{\begin{tikzpicture}[->,>=stealth,node distance=10cm, scale = 0.6]

		\node (1) at (1,6) {$\text{Income}$};
		\node (2) at (3,1) {$\text{Wine}$};
		\node (3) at (16,1) {$\text{Mortality}$};
		\node (4) at (1,-3) {$\text{U}$};
		\node (5) at (10,1) {$\text{Substance abuse}$};

		\draw [,->] (1) to [bend right=20] (2);
		\draw [,->] (1) to [bend left=20] (3);
		\draw [,->] (2) to [bend left=20] (3);
		\draw [,->] (2) to (5);
		\draw [,->] (5) to (3);
		\draw [dashed,->] (4) to [bend left=20] (2);
		\draw [dashed,->] (4) to [bend right=20] (3);

	\end{tikzpicture}}

\end{figure}

---

\frametitle{Review DAGs}

Key steps for building a quantitative research question:

>- \textbf{(1)} Draw a conceptual model to guide your research question.

>- \textbf{(2)} Read the literature. What do we know?

>- \textbf{(3)} Refine conceptual model into a DAG (this may include drawing the "causal field"). 

>- \textbf{(4)} Think systematically and explicitly about the potential effect of bias. 

---

\frametitle{DAGs}

Let's workshop a DAG on the board for someone's research project. 

---

\frametitle{Causal Inference}

>- Causal inference involves counterfactuals.

>- What is a _counterfactual_?

---

\frametitle{Potential outcomes}

>- Key assumption: “each individual in the population of interest has a potential outcome under each treatment state, even though each individual can be observed in only one treatment state at any given point in time.”

>- “The fundamental problem of causal inference”

---

\frametitle{Potential outcomes}

We observe some value of $y$ for every unit $i$:

$$y_i$$

---

\frametitle{Potential outcomes}

“The fundamental problem of causal inference”

$$\delta_i=y_{i}^1-y_{i}^0$$

---

\frametitle{Potential outcomes}

“The fundamental problem of causal inference”

$$\delta_i=y_{i}^1-y_{i}^0$$

* Experiments do not overcome this problem!
* The true experiment would be the same individual at the same time. 

---

\frametitle{Potential outcomes}

```{r, include=T}
d <- data.table(Group=c('Treatment (X=1)','Control (X=0)'),
                Y1=c('Observable','Counterfactual'),
                Y0=c('Counterfactual','Observable'))
kable(d)
```

---

\frametitle{Potential outcomes}

```{r, include=T}
d <- data.table(Group=c('Treatment (X=1)','Control (X=0)'),
                Y1=c('$y_{i}^1 \\mid X=1$','$y_{i}^1 \\mid X=0$'),
                Y0=c('$y_{i}^0 \\mid X=1$','$y_{i}^0 \\mid X=0$'))
kable(d, escape = F)
```

---

\frametitle{Potential outcomes}

```{r, include=T}
d <- data.table(Group=c('Treatment (X=1)','Control (X=0)'),
                Y1=c('Observable','Counterfactual'),
                Y0=c('Counterfactual','Observable'))
kable(d)
```

* Counterfactuals exist in theory but are \textbf{unobservable.}
* \textbf{Causal inference is a prediction problem!}

---

\frametitle{Average treatment effects}

$$y_i=\beta_1{X}+\epsilon_i$$

---

\frametitle{Average treatment effects}

$$y_i=\beta_1{X}+\epsilon_i$$

$$\beta_1=\sum_n{(y_i|X=1)-(y_i|X=0)}$$

---

\frametitle{Average treatment effects}

$$y_i=\beta_1{X}+\epsilon_i$$

$$\beta_1=\sum_n{(y_i|X=1)-(y_i|X=0)}$$

* We can't predict individual treatment effects, but we can predict the \textbf{average treatment effect (ATE).}
* The regression model is one way of generating the \textbf{predictions} needed to estimate the ATE. 

---

\frametitle{Causal inference: prediction problem}

$$\text{prestige}_i=\beta_1\text{college}_i+\beta_2\text{age}_i+\beta_3\text{race}_i+\beta_4\text{sex}_i+\beta_5\text{sibs}_i+\beta_6\text{fprestige}_i$$

\tiny
```{r, echo=F}
gss <- read_parquet('https://github.com/ngraetz/soc5811/raw/refs/heads/main/data/FinalProjectData/GSS%202021/GSS2021.parquet')
gss <- as.data.table(gss)
gss[, college := ifelse(degree %in% 3:4, 1, 0)]
gss <- gss %>%
  mutate(across(c(age,race,sex), as.numeric))
for(v in c('prestg10','college','age','race','sex','sibs','papres10')) gss <- gss[!is.na(get(v)), ]
mod <- lm(prestg10~college+age+as.factor(race)+sex+sibs+papres10,data=gss)
summary(mod)
```

---

\frametitle{Causal inference: prediction problem}

\small
```{r, echo=F}
gss[, y0 := predictions(mod, variables = list(college = 0))$estimate]
gss[, y1 := predictions(mod, variables = list(college = 1))$estimate]
head(gss[,c('college','prestg10','y0','y1')])
```

---

\frametitle{Causal inference: prediction problem}

\small
```{r, echo=T}
mean(gss$y1)
mean(gss$y0)
```
---

\frametitle{Causal inference: prediction problem}

\small
```{r, echo=T}
mean(gss$y1) - mean(gss$y0)
```

---

\frametitle{Causal inference: prediction problem}

\small
```{r, echo=T}
mean(gss$y1) - mean(gss$y0)
```

\tiny
```{r, echo=T}
summary(mod)
```

---

\frametitle{Average treatment effects}

$$y_i=\beta_1{X}+\epsilon_i$$

$$\beta_1=\sum_n{(y_i|X=1)-(y_i|X=0)}$$

>- We can't predict individual treatment effects, but we can predict the \textbf{average treatment effect (ATE).} The regression model is one way of generating the \textbf{predictions} needed to estimate the ATE. 
>- But when can these \textbf{predictions} be interpreted as \textbf{potential outcomes?}

---

\frametitle{Average treatment effects}

\textbf{Independence assumption}

$$(Y^1,Y^0,\delta) \independent  X$$

* Also referred to as exchangeability, no confounding, etc.

---

\frametitle{Average treatment effects}

```{r, include=T}
data.table(Group=c('Treatment (X=1)','Control (X=0)'),
           Y1=c('Observable','Counterfactual'),
           Y0=c('Counterfactual','Observable')) %>% 
kable() 
```

$$\text{ATE}=E[y_{i}^1]-E[y_{i}^0]$$

---

\frametitle{Average treatment effects}

```{r, include=T}
data.table(Group=c('Treatment (X=1)','Control (X=0)'),
           Y1=c('Observable','Counterfactual'),
           Y0=c('Counterfactual','Observable')) %>% 
kable() %>% 
row_spec(1, bold = TRUE, background = "lightgray") # Bolds rows 2 and 4, and adds a light gray background)
```

$$\text{ATE}=E[y_{i}^1]-E[y_{i}^0]$$

$$\textbf{ATT}=E[y_{i}^1|X=1]-E[y_{i}^0|X=1]$$

---

\frametitle{Average treatment effects}

```{r, include=T}
data.table(Group=c('Treatment (X=1)','Control (X=0)'),
           Y1=c('Observable','Counterfactual'),
           Y0=c('Counterfactual','Observable')) %>% 
kable() %>% 
row_spec(2, bold = TRUE, background = "lightgray") # Bolds rows 2 and 4, and adds a light gray background)
```

$$\text{ATE}=E[y_{i}^1]-E[y_{i}^0]$$

$$\text{ATT}=E[y_{i}^1|X=1]-E[y_{i}^0|X=1]$$

$$\textbf{ATC}=E[y_{i}^1|X=0]-E[y_{i}^0|X=0]$$

---

\frametitle{Average treatment effects}

```{r, include=T}
data.table(Group=c('Treatment (X=1)','Control (X=0)'),
           Y1=c('Observable','Counterfactual'),
           Y0=c('Counterfactual','Observable')) %>% 
kable() 
```

$$\text{ATE}=E[y_{i}^1]-E[y_{i}^0]$$

$$\text{ATT}=E[y_{i}^1|X=1]-E[y_{i}^0|X=1]$$

$$\text{ATC}=E[y_{i}^1|X=0]-E[y_{i}^0|X=0]$$

When will all three of these be identical?

---

\frametitle{Education and earnings}

