---
title: "SOC-5811 Week 10: Panel Data"
author: "Nick Graetz"
date: "11/3/2025"
output: 
    beamer_presentation:
        incremental: false
        includes: 
            in_header: Minnesota.tex
---

```{r setup, include=FALSE}
library(AER)
library(data.table)
library(knitr)
library(kableExtra)
library(arrow)
library(haven)
library(ggplot2)
library(marginaleffects)
library(tidyverse)
library(fixest)
knitr::opts_chunk$set(
 fig.width = 6,
 fig.height = 4,
 warning = FALSE, 
 message = FALSE,
 out.width="80%"
)
dropbox <- 'C:/Users/ncgra/Dropbox/'
fig_dir <- paste0(dropbox, 'Minnesota/repos/soc5811/figures/')
```

---

\frametitle{Panel data}

>- What is "panel data?" 

>- \textbf{Panel data:} A data set where we observe the same units (e.g., individuals, firms, countries, schools) over more than one time period. 

>- Walk through data examples: when is longitudinal data panel data? Repeated observations of the same variable.

---

\frametitle{Panel data}

Basic linear regression model: 

$$y_i=\beta_0+\beta_1X_i$$

Linear regression model with panel data:

$$y_{i,t}=\beta_0+\beta_1X_{i,t}$$

---

\frametitle{Panel data}

>- Let's walk through an example with data: traffic deaths and beer taxes by state/year. 

---

\frametitle{Panel data}

\tiny
```{r, echo=F}
data("Fatalities")
Fatalities <- as.tibble(Fatalities)
```

\tiny
```{r, echo=T}
dim(Fatalities)
Fatalities %>% select(state) %>% table()
Fatalities %>% select(year) %>% table()
```

---

\frametitle{Panel data}

>- This is called "balanced" panel data. Why? What would be "unbalanced" panel data?

>- What shape is this dataset with respect to time?

---

\frametitle{Panel data}

\tiny
```{r, echo=T}
head(Fatalities)
```

---

\frametitle{Panel data}

$$\text{Rate}_{i,t}=\beta_0+\beta_1\text{Tax}_{i,t}$$

---

\frametitle{Panel data}

All years (pooled model): 

```{r, echo=F}
Fatalities <- Fatalities %>% 
  mutate(rate = fatal / pop * 100000)
ggplot(data=Fatalities,
       aes(x=beertax,y=rate)) + 
  geom_point() + 
  geom_smooth(method='lm') + 
  theme_bw()
```

---

\frametitle{Panel data}

\tiny
```{r, echo=T}
mod <- lm(rate ~ beertax, data=Fatalities)
summary(mod)
```

---

\frametitle{Panel data}

We could just think of panel data as several samples and several regression models:

$$\text{Rate}_{i,1988}=\beta_0+\beta_1\text{Tax}_{i,1988}$$

$$\text{Rate}_{i,1982}=\beta_0+\beta_1\text{Tax}_{i,1982}$$

---

\frametitle{Panel data}

$$\text{Rate}_{i,1988}=\beta_0+\beta_1\text{Tax}_{i,1988}$$

$$\text{Rate}_{i,1982}=\beta_0+\beta_1\text{Tax}_{i,1982}$$

$$\text{Rate}_{i,1988} - \text{Rate}_{i,1982}=\beta_0+\beta_1(\text{Tax}_{i,1988}-\text{Tax}_{i,1982})$$
We have now made this a "within" estimator: we are predicting within-unit changes. 

---

\frametitle{Panel data}

\tiny
```{r, echo=T}
Fatalities_wide <- Fatalities %>%
  filter(year %in% c(1982,1988)) %>%
  pivot_wider(id_cols = state, 
              values_from = c(rate,beertax), 
              names_from = year) %>%
  mutate(rate_change = rate_1988 - rate_1982,
         tax_change = beertax_1988 - beertax_1982)
```

---

\frametitle{Panel data}

\tiny
```{r, echo=T}
head(Fatalities_wide)
```

---

\frametitle{Panel data}

\tiny
```{r, echo=T}
mod <- lm(rate_change ~ tax_change, data=Fatalities_wide)
summary(mod)
```

---

\frametitle{Panel data}

>- What does this remind you of that we have covered already?

>- Simpson's Paradox! What was the idea of Simpson's Paradox related to omitted variables?

>- There was some variable we were missing. If we looked _within_ penguin species, we found the opposite relationship!

>- \textbf{Intercept shifts.}

---

\frametitle{Causal inference}

>- What is the difference between Simpson's Paradox examples and what we are doing here? 

>- There is no new _observed variable_ that we are adding. 

>- So what have we done? We have accounted for _unobserved variables_.

>- But what specific kind of unobserved variables? Time-invariant

---

\frametitle{Causal inference}

>- Panel data and causal inference

$$\text{Rate}_{i,1988}=\beta_0+\beta_1\text{Tax}_{i,1988}+\beta_2 Z_i$$

$$\text{Rate}_{i,1982}=\beta_0+\beta_1\text{Tax}_{i,1982}+\beta_2 Z_i$$

$$\text{Rate}_{i,1988} - \text{Rate}_{i,1982}=\beta_0+\beta_1(\text{Tax}_{i,1988}-\text{Tax}_{i,1982})$$

---

\frametitle{Causal inference}

>- When might we run into problems?

---

\frametitle{Review}

>- Pick up here on Wednesday

---

\frametitle{Panel data}

\textbf{Panel data:} A data set where we observe the same units (e.g., individuals, firms, countries, schools) over more than one time period. 

\textbf{Linear regression model with cross-sectional data:} 

$$y_i=\beta_0+\beta_1X_i$$

\textbf{Linear regression model with panel data:}

$$y_{i,t}=\beta_0+\beta_1X_{i,t}$$

---

\frametitle{Cross-sectional data}

\begin{center}
\includegraphics[width = 0.8\textwidth]{images/penguin2.png}
\end{center}

---

\frametitle{Cross-sectional data}

\textbf{Cross-sectional} data on penguin bill length and depth. 

\tiny
```{r, echo=T}
penguin_df <- 
  palmerpenguins::penguins %>%
  na.omit()
head(penguin_df)
```

---

\frametitle{Cross-sectional data}

$$\text{Depth}_i=\beta_0+\boxed{\beta_1}\text{Length}_i$$

---

\frametitle{Cross-sectional data}

>- What variation are we using to estimate the coefficient? 

```{r, echo=F}
penguin_df <- 
  palmerpenguins::penguins %>%
  na.omit()
mod <- lm(bill_depth_mm ~ bill_length_mm + species, data = penguin_df)
predictions <- tibble(depth_preds = predictions(mod)$estimate)
penguin_df <- penguin_df %>%
  bind_cols(., predictions)
ggplot(data=penguin_df,
       aes(x=bill_length_mm, y=bill_depth_mm)) + 
  geom_point(shape=21,size=3,fill='lightgrey') + 
  geom_smooth(method='lm',se=F,color='red') + 
  labs(x="Length", y="Depth", fill='Species') +
  theme_bw()
```

---

\frametitle{Cross-sectional data}

>- \textbf{Between-unit variation}.

```{r, echo=F}
ggplot(data=penguin_df,
       aes(x=bill_length_mm, y=bill_depth_mm)) + 
  geom_point(shape=21,size=3,fill='lightgrey') + 
  geom_smooth(method='lm',se=F,color='red') + 
  labs(x="Length", y="Depth", fill='Species') +
  theme_bw()
```

---

\frametitle{Cross-sectional data}

What variation are we using once we control for species?

$$\text{Depth}_i=\beta_0+\boxed{\beta_1}\text{Length}_i+\beta_2\text{Species}_i$$

---

\frametitle{Cross-sectional data}

\small
```{r, echo=T}
penguin_df <- penguin_df %>% 
  group_by(species) %>% 
  mutate(bill_depth_mm_demeaned = 
           bill_depth_mm - mean(bill_depth_mm),
         bill_length_mm_demeaned = 
           bill_length_mm - mean(bill_length_mm))
```

---

\frametitle{Cross-sectional data}

\tiny
```{r, echo=F}
ggplot(data=penguin_df,
       aes(x=bill_length_mm_demeaned, y=bill_depth_mm_demeaned, fill=species)) + 
  geom_point(shape=21,size=3) + 
  geom_smooth(aes(x=bill_length_mm_demeaned, y=bill_depth_mm_demeaned), method='lm',se=F,size=1, color='red', inherit.aes = F) +
  scale_fill_viridis_d() + 
  labs(x="Length", y="Depth", fill='Species') +
  theme_bw()
```

---

\frametitle{Cross-sectional data}

>- What variation are we using to estimate this coefficient?
>- Because we controlled for species, we are now using within-species variation. But with respect to unit, we are still using \textbf{between-unit} variation to estimate the coefficient.

---

\frametitle{Cross-sectional data}

>- Intercept shift by species.

```{r, echo=F}
ggplot(data=penguin_df,
       aes(x=bill_length_mm, y=bill_depth_mm, fill=species)) + 
  geom_point(shape=21,size=3) + 
  geom_line(aes(y=depth_preds),size=1,color='red') + 
  scale_fill_viridis_d() + 
  labs(x="Length", y="Depth", fill='Species') +
  theme_bw()
```

---

\frametitle{Panel data}

\textbf{Panel data} on traffic mortality rate and beer taxes. 

\tiny
```{r, echo=T}
head(Fatalities)
```

---

\frametitle{Panel data}

$$\text{Rate}_{i,t}=\beta_0+\boxed{\beta_1}\text{Tax}_{i,t}$$

---

\frametitle{Panel data}

>- What variation are we using to estimate the coefficient? 

```{r, echo=F}
mod <- lm(rate~beertax+state, data=Fatalities)
predictions <- tibble(rate_preds = predictions(mod)$estimate)
Fatalities <- Fatalities %>%
  bind_cols(., predictions)
ggplot(data=Fatalities,
       aes(x=beertax, y=rate)) + 
  geom_point(shape=21,size=3,fill='lightgrey') + 
  geom_smooth(method='lm',se=F,color='red') + 
  labs(x="Beer tax", y="Traffic mortality rate per 100,000") +
  theme_bw()
```

---

\frametitle{Panel data}

>- \textbf{Between-unit/time variation}.

```{r, echo=F}
ggplot(data=Fatalities,
       aes(x=beertax, y=rate)) + 
  geom_point(shape=21,size=3,fill='lightgrey') + 
  geom_smooth(method='lm',se=F,color='red') + 
  labs(x="Beer tax", y="Traffic mortality rate per 100,000") +
  theme_bw()
```

---

\frametitle{Panel data}

What variation are we using once we control for state?

$$\text{Rate}_{i,t}=\beta_0+\boxed{\beta_1}\text{Tax}_{i,t}+\beta_2\text{State}_i$$

---

\frametitle{Panel data}

\small
```{r, echo=T}
Fatalities <- Fatalities %>%
  group_by(state) %>% 
  mutate(rate_demeaned = rate - mean(rate),
         tax_demeaned = beertax - mean(beertax))
mod <- lm(rate_demeaned ~ tax_demeaned, data=Fatalities)
```

---

\frametitle{Panel data}

```{r, echo=F}
predictions <- tibble(rate_preds2 = predictions(mod)$estimate)
Fatalities <- Fatalities %>%
  bind_cols(., predictions)
ggplot(data=Fatalities,
       aes(x=tax_demeaned, y=rate_demeaned)) + 
  geom_point(shape=21,size=3,fill='lightgrey') + 
  geom_line(aes(y=rate_preds2),size=1,color='red') + 
  scale_fill_viridis_d() + 
  labs(x="Beer tax (DEMEANED)", y="Traffic mortality rate per 100,000 (DEMEANED)") +
  theme_bw()
```

---

\frametitle{Panel data}

>- What variation are we using to estimate this coefficient?
>- Because we controlled for state, we are now using \textbf{within-unit} variation over _time_ to estimate slope. 

---

\frametitle{Panel data}

>- Intercept shift by _unit_ (state).

```{r, echo=F}
ggplot(data=Fatalities,
       aes(x=beertax, y=rate, group=state)) + 
  geom_point(shape=21,size=3,fill='lightgrey') + 
  geom_line(aes(y=rate_preds),size=1,color='red') + 
  scale_fill_viridis_d() + 
  labs(x="Beer tax", y="Traffic mortality rate per 100,000") +
  theme_bw()
```

---

\frametitle{Panel data}

\small
```{r, echo=T}
mod <- lm(rate ~ beertax + state, data=Fatalities)
```

---

\frametitle{Panel data}

\tiny
```{r, echo=F}
summary(mod)
```

---

\frametitle{Panel data}

\small
```{r, echo=T}
Fatalities <- Fatalities %>%
  group_by(state) %>% 
  mutate(rate_demeaned = rate - mean(rate),
         tax_demeaned = beertax - mean(beertax))
mod <- lm(rate_demeaned ~ tax_demeaned, data=Fatalities)
```

---

\frametitle{Panel data}

\tiny
```{r, echo=F}
summary(mod)
```

---

\frametitle{Intercept shifts}

\textbf{Penguin-level} analysis of bill length and depth ($\beta_1$):

$$\text{depth}_i=\beta_0+\beta_1\text{length}_i+\beta_2\text{species}_i$$

\textbf{State-level} analysis of traffic deaths and taxes ($\beta_1$): 

$$\text{rate}_{i,t}=\beta_0+\beta_1\text{tax}_{i,t}+\beta_2\text{state}_i$$

>- With panel data, we're basically doing the same sort of model as before - but now we are controlling for the _unit_ of analysis itself. 
>- \textbf{Next week:} implications for _estimation_ (i.e., many units) and _uncertainty_ (i.e., correlated errors). 

---

\frametitle{Causal inference}

>- What might panel data and within-unit variation help us do from a _causal_ perspective?

>- Let's think about trying to estimate a causal effect using \textbf{between-unit} variation, as we discussed last week.

---

\frametitle{Average treatment effects}

$$y_i=\beta_1{X}+\epsilon_i$$

$$\beta_1=\sum_n{(y_i|X=1)-(y_i|X=0)}$$

>- But when can these \textbf{predictions} be interpreted as \textbf{potential outcomes?}

>- We can't predict individual treatment effects, but we can predict the \textbf{average treatment effect (ATE).} The regression model is one way of generating the \textbf{predictions} needed to estimate the ATE. 

---

\frametitle{Average treatment effects}

\textbf{Independence assumption}

$$(Y^1,Y^0,\delta) \independent  X$$

>- Also referred to as assuming exchangeability, no unobserved confounding, etc.

>- What do we assume is exchangeable?

---

\frametitle{Causal inference}

\begin{center}
\includegraphics[width = 1\textwidth]{images/matching.png}
\end{center}

---

\frametitle{Causal inference}

>- Implication: there is always unobserved confounding, which might be really consequential when estimating effects using \textbf{between-unit} variation from observational data.

\begin{center}
\includegraphics[width = 1\textwidth]{images/matching.png}
\end{center}

---

\frametitle{Causal inference}

>- But what if we had longitudinal panel data? What if we can estimate effects using \textbf{within-unit} variation? 

\begin{center}
\includegraphics[width = 1\textwidth]{images/matching.png}
\end{center}

---

\frametitle{Causal inference}

>- Panel data and causal inference

$$\text{Rate}_{i,1988}=\beta_0+\beta_1\text{Tax}_{i,1988}+\beta_2 Z_i$$

$$\text{Rate}_{i,1982}=\beta_0+\beta_1\text{Tax}_{i,1982}+\beta_2 Z_i$$

$$\text{Rate}_{i,1988} - \text{Rate}_{i,1982}=\beta_0+\beta_1(\text{Tax}_{i,1988}-\text{Tax}_{i,1982})$$

---

\frametitle{Causal inference}

>- Panel data and causal inference: remember the relationship to demeaning! We are removing the \textbf{unit-specific means}; in other words, all \textbf{time-invariant} characteristics of units. 

$$\text{Rate}_{i,t} - \overline{\text{Rate}_i}=\beta_0+\beta_1(\text{Tax}_{i,t}-\overline{\text{Tax}_i})$$

>- Why is this magical? Why is this different from the penguin example where we had to _observe_ and control for species? 

>- Because we are removing lots of characteristics that are \textbf{unobserved}!

---

\frametitle{Causal inference}

\begin{center}
\includegraphics[width = 1\textwidth]{images/matching.png}
\end{center}

---

\frametitle{Causal inference}

\textbf{Caveats:}

>- \textbf{(1)} Unit-level fixed effects cannot address \textbf{time-variant} unobserved confounding. 
>   - Even though we are only using within-unit variation, it's still possible that there was _something else_ changing over time within units that caused the outcome. 

>- \textbf{(2)} Unit-level fixed effects can lead to substantial bias if there is strong reverse causality. 
>   - For example, if $Y \longrightarrow X$, then unit-level fixed effects will over-control the mediating pathway by which part of the total effect of $Y \longrightarrow X$ over the period operates through $X_{i,1} \longrightarrow Y_{i,2} \longrightarrow X_{i,3} \longrightarrow Y_{i,4}$

---

\frametitle{Causal inference}

\textbf{Caveats:}

>- \textbf{(3)} Including fixed effects isn't just a "better" way to estimate the same estimand; it changes the estimand. 
>   - Examining how a target variable predicts \textbf{within-unit} variation in an outcome, while very useful for isolating causal effects, is not the same thing as examining how a target variable predicts \textbf{between-unit} variation, which is often more relevant for descriptive questions. 
>   - What predicts differences in the average maternal mortality rate across hospitals? Even if I have panel data on hospitals over time, I may not want to include a hospital-level fixed effect. I am more interested in comparing hospitals than I am in comparing changes over time within hospitals. 

---

\frametitle{Causal inference}

\textbf{Caveats:}

* \textbf{(1)} Unit-level fixed effects cannot address \textbf{time-variant} unobserved confounding. 

* \textbf{(2)} Unit-level fixed effects can lead to substantial bias if there is strong reverse causality. 

* \textbf{(3)} Including fixed effects isn't just a "better" way to estimate the same comparison; it _changes_ the comparison.

If you are interested in further developing your causal inference skills, I teach a course called \textbf{SOC8890: Applied Causal Inference.}

