---
title: "SOC-5811 Week 10: Panel Data"
author: "Nick Graetz"
date: "11/3/2025"
output: 
    beamer_presentation:
        incremental: false
        includes: 
            in_header: Minnesota.tex
---

```{r setup, include=FALSE}
library(AER)
library(data.table)
library(knitr)
library(kableExtra)
library(arrow)
library(haven)
library(ggplot2)
library(marginaleffects)
library(tidyverse)
knitr::opts_chunk$set(
 fig.width = 6,
 fig.height = 4,
 warning = FALSE, 
 message = FALSE,
 out.width="80%"
)
dropbox <- 'C:/Users/ncgra/Dropbox/'
fig_dir <- paste0(dropbox, 'Minnesota/repos/soc5811/figures/')
```

---

\frametitle{Panel data}

>- What is "panel data?" 

>- \textbf{Panel data:} A data set where we observe the same units (e.g., individuals, firms, countries, schools) over more than one time period. 

>- Walk through data examples: when is longitudinal data panel data? Repeated observations of the same variable.

---

\frametitle{Panel data}

Basic linear regression model: 

$$y_i=\beta_0+\beta_1X_i$$

Linear regression model with panel data:

$$y_{i,t}=\beta_0+\beta_1X_{i,t}$$

---

\frametitle{Panel data}

>- Let's walk through an example with data: traffic deaths and beer taxes by state/year. 

---

\frametitle{Panel data}

\tiny
```{r, echo=F}
data("Fatalities")
Fatalities <- as.tibble(Fatalities)
```

\tiny
```{r, echo=T}
dim(Fatalities)
Fatalities %>% select(state) %>% table()
Fatalities %>% select(year) %>% table()
```

---

\frametitle{Panel data}

>- This is called "balanced" panel data. Why? What would be "unbalanced" panel data?

>- What shape is this dataset with respect to time?

---

\frametitle{Panel data}

\tiny
```{r, echo=T}
head(Fatalities)
```

---

\frametitle{Panel data}

$$\text{Rate}_{i,t}=\beta_0+\beta_1\text{Tax}_{i,t}$$

---

\frametitle{Panel data}

All years (pooled model): 

```{r, echo=F}
Fatalities <- Fatalities %>% 
  mutate(rate = fatal / pop * 100000)
ggplot(data=Fatalities,
       aes(x=beertax,y=rate)) + 
  geom_point() + 
  geom_smooth(method='lm') + 
  theme_bw()
```

---

\frametitle{Panel data}

\tiny
```{r, echo=T}
mod <- lm(rate ~ beertax, data=Fatalities)
summary(mod)
```

---

\frametitle{Panel data}

We could just think of panel data as several samples and several regression models:

$$\text{Rate}_{i,1988}=\beta_0+\beta_1\text{Tax}_{i,1988}$$

$$\text{Rate}_{i,1982}=\beta_0+\beta_1\text{Tax}_{i,1982}$$

---

\frametitle{Panel data}

$$\text{Rate}_{i,1988}=\beta_0+\beta_1\text{Tax}_{i,1988}$$

$$\text{Rate}_{i,1982}=\beta_0+\beta_1\text{Tax}_{i,1982}$$

$$\text{Rate}_{i,1988} - \text{Rate}_{i,1982}=\beta_0+\beta_1(\text{Tax}_{i,1988}-\text{Tax}_{i,1982})$$
We have now made this a "within" estimator: we are predicting within-unit changes. 

---

\frametitle{Panel data}

\tiny
```{r, echo=T}
Fatalities_wide <- Fatalities %>%
  filter(year %in% c(1982,1988)) %>%
  pivot_wider(id_cols = state, 
              values_from = c(rate,beertax), 
              names_from = year) %>%
  mutate(rate_change = rate_1988 - rate_1982,
         tax_change = beertax_1988 - beertax_1982)
```

---

\frametitle{Panel data}

\tiny
```{r, echo=T}
head(Fatalities_wide)
```

---

\frametitle{Panel data}

\tiny
```{r, echo=T}
mod <- lm(rate_change ~ tax_change, data=Fatalities_wide)
summary(mod)
```

---

\frametitle{Panel data}

>- What does this remind you of that we have covered already?

>- Simpson's Paradox! What was the idea of Simpson's Paradox related to omitted variables?

>- There was some variable we were missing. If we looked _within_ penguin species, we found the opposite relationship!

>- \textbf{Intercept shifts.}

---

\frametitle{Causal inference}

>- What is the difference between Simpson's Paradox examples and what we are doing here? 

>- There is no new _observed variable_ that we are adding. 

>- So what have we done? We have accounted for _unobserved variables_.

>- But what specific kind of unobserved variables? Time-invariant

---

\frametitle{Causal inference}

>- Panel data and causal inference

$$\text{Rate}_{i,1988}=\beta_0+\beta_1\text{Tax}_{i,1988}+\beta_2 Z_i$$

$$\text{Rate}_{i,1982}=\beta_0+\beta_1\text{Tax}_{i,1982}+\beta_2 Z_i$$

$$\text{Rate}_{i,1988} - \text{Rate}_{i,1982}=\beta_0+\beta_1(\text{Tax}_{i,1988}-\text{Tax}_{i,1982})$$

---

\frametitle{Causal inference}

>- When might we run into problems?
