---
title: "SOC-5811 Week 11: Panel Data"
author: "Nick Graetz"
date: "11/10/2025"
output: 
    beamer_presentation:
        incremental: false
        includes: 
            in_header: Minnesota.tex
---

```{r setup, include=FALSE}
library(AER)
library(data.table)
library(knitr)
library(kableExtra)
library(arrow)
library(haven)
library(ggplot2)
library(marginaleffects)
library(tidyverse)
library(fixest)
knitr::opts_chunk$set(
 fig.width = 6,
 fig.height = 4,
 warning = FALSE, 
 message = FALSE,
 out.width="80%"
)
dropbox <- 'C:/Users/ncgra/Dropbox/'
fig_dir <- paste0(dropbox, 'Minnesota/repos/soc5811/figures/')
```

---

\frametitle{Review: Interactions}

$$\text{PolViews}_i=\beta_0+\beta_1\text{Religion}_i+\beta_2\text{Age}_i$$

---

\frametitle{Review: Interactions}

$$\text{PolViews}_i=\beta_0+\beta_1\text{Religion}_i+\beta_2\text{Age}_i+\beta_3(\text{Religion}_i\times\text{Age}_i)$$

---

\frametitle{Review: Interactions}

$$\text{PolViews}_i=\beta_0+\boxed{\beta_1}\text{Religion}_i+\beta_2\text{Age}_i+\beta_3(\text{Religion}_i\times\text{Age}_i)$$

---

\frametitle{Review}

>- What do "fixed effects" do in linear regression models fit to panel data?

$$y_{i,t}=\beta_0+\beta_1X_{i,t}+\alpha_i$$

>- Basically the same thing as controlling for a categorical variable, but the "category" is units.
>   - Intercept shift by unit.
>   - Isolates our target comparison to be based on average _within-unit_ variation (i.e., variation over time). 
>   - From a causal perspective, eliminates all \textbf{time-invariant} confounders.

---

\frametitle{Review}

* \textbf{(1)} Unit-level fixed effects cannot address \textbf{time-variant} unobserved confounding. 

* \textbf{(2)} Unit-level fixed effects can lead to substantial bias if there is strong reverse causality. 

* \textbf{(3)} Including fixed effects isn't just a "better" way to estimate the same comparison; it _changes_ the comparison.

---

\frametitle{Review}

>- When do I want to estimate coefficients using \textbf{between-unit} differences vs. \textbf{within-unit} differences? 

>- If you care about estimating \textbf{causal effects} (or something you want to generally interpret that way), you probably want to use within-unit differences.

>- If you care about estimating \textbf{descriptive disparities}, you probably want to use between-unit differences. 

---

\frametitle{Between vs. within}

$$i=\text{hospital}$$
$$t=\text{year}$$

$$\text{Physicians}_{i,t}=\beta_0+\beta_1\text{SafetyNet}_{i,t}+\beta_2Z$$

$$\text{Physicians}_{i,t}=\beta_0+\beta_1\text{SafetyNet}_{i,t}+\beta_2Z+\alpha_i$$

---

\frametitle{Between vs. within}

$$\text{Physicians}_{i,t}=\beta_0+\boxed{\beta_1}\text{SafetyNet}_{i,t}+\beta_2Z$$

$$\text{Physicians}_{i,t}=\beta_0+\boxed{\beta_1}\text{SafetyNet}_{i,t}+\beta_2Z+\alpha_i$$

---

\frametitle{Between vs. within}

>- \textbf{Research question:} Do safety net hospitals have different average physicians per patient, controlling for $Z$? 
>- $\beta_1$: Across all hospitals, safety net hospitals have $\beta_1$ more/fewer physicians per capita.
>- Descriptive comparison with implications for stratification/inequality. 

$$\text{Physicians}_{i,t}=\beta_0+\boxed{\beta_1}\text{SafetyNet}_{i,t}+\beta_2Z$$

---

\frametitle{Between vs. within}

>- \textbf{Research question:} As the patient population changes within hospitals, does average physicians per patient change? 
>- Causal comparison with implications for the effect of patient population on work force within hospitals. When hospitals become safety net hospitals, do they see reductions in work force? 

$$\text{Physicians}_{i,t}=\beta_0+\boxed{\beta_1}\text{SafetyNet}_{i,t}+\beta_2Z+\alpha_i$$

---

\frametitle{Between vs. within}

* \textbf{Research question:} Do safety net hospitals have different average physicians per patient, controlling for $Z$? 

* \textbf{Research question:} As the patient population changes within hospitals, does average physicians per patient change?  

* \textbf{Both answer interesting but different research questions.}

---

\frametitle{Estimation}

How to fit fixed effects models: 

>- OLS estimator: Add unit as a categorical variable in lm().

>- OLS estimator: Manually demean and then lm() with demeaned data.

>- Fixed effects estimators (modified OLS): \textbf{feols().}

\tiny
```{r, echo=F}
data("Fatalities")
Fatalities <- as.tibble(Fatalities)
Fatalities <- Fatalities %>% 
  mutate(rate = fatal / pop * 100000)
Fatalities <- Fatalities %>%
  group_by(state) %>% 
  mutate(rate_demeaned = rate - mean(rate),
         tax_demeaned = beertax - mean(beertax))
```

---

\frametitle{Estimation}

\tiny
```{r, echo=T}
summary(lm(rate ~ beertax + state, data=Fatalities))
```

---

\frametitle{Estimation}

\tiny
```{r, echo=T}
summary(lm(rate_demeaned ~ tax_demeaned, data=Fatalities))
```

---

\small
```{r, echo=T}
summary(feols(rate ~ beertax | state, data=Fatalities))
```

---

\frametitle{Estimation}

How to fit fixed effects models: 

>- OLS estimator: Add unit as a categorical variable in lm().
>   - Very slow/clunky if you have thousands of units. 

>- OLS estimator: Manually demean and then lm() with demeaned data.
>   - Very complex to do correctly by hand if you have many variables in your model. 

>- Fixed effects estimators (modified OLS): \textbf{feols().}
>   - Very fast and efficient; the tradeoff is that you don't estimate standard errors on your fixed effects themselves, but typically this does not matter for your target comparison. 

---

\frametitle{Uncertainty}

>- It is tricky to get the standard errors correct with panel data. Why?

>- Remember that in the conventional regression model, we assume errors are \textbf{independent.} In sampling terms, we assume every sample from the population model is independent of any other sample.

>- Think about flipping a coin 100 times. When we estimate the standard error of the random variable $X$ = number of heads in 100 flips, we assume each flip is unrelated to (or independent from) the other flips.

>- Solution: \textbf{clustered standard errors.}

---

\frametitle{Uncertainty}

\small
\textbf{$\beta_1$ from our basic OLS model with unit as a categorical variable:}

```{r, echo=F}
mod1 <- summary(lm(rate ~ beertax + state, data=Fatalities))
mod1[['coefficients']][2,]
```
\textbf{$\beta_1$ from our fixed effects model:}

```{r, echo=F}
mod2 <- summary(feols(rate ~ beertax | state, data=Fatalities))
summary(mod2)
```

---

\frametitle{Review}

Pick up here on Wednesday.

---

\frametitle{Interactions Review}

\textbf{Categorical} $\times$ \textbf{Categorical}

$$\text{MPG}_i=\beta_0+\beta_1\text{E}_i+\beta_2\text{T}_i+\beta_3(\text{E}_i\times\text{T}_i)$$

---

\textbf{Categorical} $\times$ \textbf{Categorical}

$$\text{MPG}_i=\beta_0+\beta_1\text{E}_i^{\text{v-shaped}}+\beta_2\text{T}_i^{\text{manual}}+\beta_3(\text{E}_i^{\text{v-shaped}}\times\text{T}_i^{\text{manual}})$$

---

\textbf{Categorical} $\times$ \textbf{Categorical}

$$\text{MPG}_i=\beta_0+\beta_1\text{E}_i^{\text{v-shaped}}+\beta_2\text{T}_i^{\text{manual}}+\beta_3(\text{E}_i^{\text{v-shaped}}\times\text{T}_i^{\text{manual}})$$

\tiny
```{r, echo=F} 
data(mtcars)
mtcars <- mtcars %>%
  mutate(engine_type=recode(vs,`0`='v-shaped',`1`='straight'),
         transmission=recode(am,`0`='automatic',`1`='manual'))
mod <- lm(mpg~engine_type*transmission, data=mtcars)
summary(mod)
```

---

\textbf{Categorical} $\times$ \textbf{Categorical}

\small
$$\text{MPG}_i=(20.7)+(-5.7)\text{E}_i^{\text{v-shaped}}+(7.6)\text{T}_i^{\text{manual}}+(-2.9)(\text{E}_i^{\text{v-shaped}}\times\text{T}_i^{\text{manual}})$$

\tiny
```{r, echo=F} 
data(mtcars)
mtcars <- mtcars %>%
  mutate(engine_type=recode(vs,`0`='v-shaped',`1`='straight'),
         transmission=recode(am,`0`='automatic',`1`='manual'))
mod <- lm(mpg~engine_type*transmission, data=mtcars)
summary(mod)
```

---

\textbf{Categorical} $\times$ \textbf{Categorical}

\small
$$\text{MPG}_i=(20.7)+(-5.7)\text{E}_i^{\text{v-shaped}}+(7.6)\text{T}_i^{\text{manual}}+(-2.9)(\text{E}_i^{\text{v-shaped}}\times\text{T}_i^{\text{manual}})$$

```{r, echo=F} 
preds <- predictions(mod, 
                     condition = c('engine_type','transmission'))
preds <- as.data.table(preds)
preds <- unique(preds[,c('engine_type','transmission','estimate')])
ggplot(data=preds,
       aes(x=engine_type,y=estimate,fill=transmission)) + 
  geom_point(size=5,shape=21) + 
  labs(y='MPG',x='Engine Type',fill='Transmission Type') + 
  theme_bw()
```

---

\textbf{Categorical} $\times$ \textbf{Categorical}

\small
$$\text{MPG}_i=(20.7)+(-5.7)\text{E}_i^{\text{v-shaped}}+(7.6)\text{T}_i^{\text{manual}}+(-2.9)(\text{E}_i^{\text{v-shaped}}\times\text{T}_i^{\text{manual}})$$
```{r, echo=F} 
ggplot(data=preds,
       aes(x=engine_type,y=estimate,fill=transmission)) + 
  geom_point(size=5,shape=21) + 
  geom_linerange(x='straight',ymin=15.05000,ymax=20.74286,size=1) + 
  labs(y='MPG',x='Engine Type',fill='Transmission Type') + 
  theme_bw()
```

---

\textbf{Categorical} $\times$ \textbf{Categorical}

\small
$$\text{MPG}_i=(20.7)+(-5.7)\text{E}_i^{\text{v-shaped}}+(7.6)\text{T}_i^{\text{manual}}+(-2.9)(\text{E}_i^{\text{v-shaped}}\times\text{T}_i^{\text{manual}})$$
```{r, echo=F} 
ggplot(data=preds,
       aes(x=engine_type,y=estimate,fill=transmission)) + 
  geom_point(size=5,shape=21) + 
  geom_linerange(x='straight',ymin=20.74286,ymax=28.37143,size=1) + 
  labs(y='MPG',x='Engine Type',fill='Transmission Type') + 
  theme_bw()
```

---

\textbf{Categorical} $\times$ \textbf{Categorical}

\small
$$\text{MPG}_i=(20.7)+(-5.7)\text{E}_i^{\text{v-shaped}}+(7.6)\text{T}_i^{\text{manual}}+(-2.9)(\text{E}_i^{\text{v-shaped}}\times\text{T}_i^{\text{manual}})$$
```{r, echo=F} 
ggplot(data=preds,
       aes(x=engine_type,y=estimate,fill=transmission)) + 
  geom_point(size=5,shape=21) + 
  geom_linerange(x='straight',ymin=20.74286,ymax=28.37143,size=1) + 
  geom_linerange(x='v-shaped',ymin=15.05000,ymax=19.75000,size=1) + 
  labs(y='MPG',x='Engine Type',fill='Transmission Type') + 
  theme_bw()
```

---

\textbf{Categorical} $\times$ \textbf{Categorical}

\small
$$\text{MPG}_i=(20.7)+(-5.7)\text{E}_i^{\text{v-shaped}}+(7.6)\text{T}_i^{\text{manual}}+(-2.9)(\text{E}_i^{\text{v-shaped}}\times\text{T}_i^{\text{manual}})$$
```{r, echo=F} 
ggplot(data=preds,
       aes(x=engine_type,y=estimate,fill=transmission)) + 
  geom_point(size=5,shape=21) + 
  geom_linerange(x='straight',ymin=15.05000,ymax=20.74286,size=1) + 
  geom_linerange(x='v-shaped',ymin=19.75000,ymax=28.37143,size=1) + 
  labs(y='MPG',x='Engine Type',fill='Transmission Type') + 
  theme_bw()
```

---

\textbf{Categorical} $\times$ \textbf{Categorical}

\small
$$\text{MPG}_i=(20.7)+(-5.7)\text{E}_i^{\text{v-shaped}}+(7.6)\text{T}_i^{\text{manual}}+(-2.9)(\text{E}_i^{\text{v-shaped}}\times\text{T}_i^{\text{manual}})$$

* $\beta_1$: Across all cars at the reference level for Transmission (Transmission=automatic), v-shaped Engines are associated with -5.7 lower MPG. 

* $\beta_2$: Across all cars at the reference level for Engine (Engine=straight), manual Transmissions are associated with 7.6 higher MPG. 

* $\beta_3$: The association between v-shaped Engines and MPG for Transmission=automatic (-5.7) is -2.9 lower for Transmission=manual. The association between manual Transmissions and MPG for Engine=straight (7.6) is -2.9 lower for Engine=v-shaped. 

---

\frametitle{Interactions Review}

\textbf{Continuous} $\times$ \textbf{Categorical}

$$\text{MPG}_i=\beta_0+\beta_1\text{H}_i+\beta_2\text{C}_i+\beta_3(\text{H}_i\times\text{C}_i)$$

---

\textbf{Continuous} $\times$ \textbf{Categorical}

\small
$$\text{MPG}_i=\beta_0+\beta_1\text{H}_i+\beta_2\text{C}_i^{\text{6}}+\beta_3\text{C}^8+\beta_4(\text{H}_i\times\text{C}_i^6)+\beta_5(\text{H}_i\times\text{C}_i^8)$$

---

\textbf{Continuous} $\times$ \textbf{Categorical}

\small
$$\text{MPG}_i=\beta_0+\beta_1\text{H}_i+\beta_2\text{C}_i^{\text{6}}+\beta_3\text{C}^8+\beta_4(\text{H}_i\times\text{C}_i^6)+\beta_5(\text{H}_i\times\text{C}_i^8)$$

\tiny
```{r, echo=F} 
mod <- lm(mpg~hp*as.factor(cyl), data=mtcars)
summary(mod)
```

---

\textbf{Continuous} $\times$ \textbf{Categorical}

\small
$$\text{MPG}_i=(36)+(-0.1)\text{H}_i+(-15)\text{C}_i^{\text{6}}+(-18)\text{C}^8+(0.1)(\text{H}_i\times\text{C}_i^6)+(0.1)(\text{H}_i\times\text{C}_i^8)$$

\tiny
```{r, echo=F} 
mod <- lm(mpg~hp*as.factor(cyl), data=mtcars)
summary(mod)
```

---

\textbf{Continuous} $\times$ \textbf{Categorical}

\small
$$\text{MPG}_i=(36)+(-0.1)\text{H}_i+(-15)\text{C}_i^{\text{6}}+(-18)\text{C}^8+(0.1)(\text{H}_i\times\text{C}_i^6)+(0.1)(\text{H}_i\times\text{C}_i^8)$$

```{r, echo=F} 
preds <- predictions(mod, 
                     condition = c('hp','cyl'), 
                     newdata = datagrid(hp=50:350,cyl=c(4,6,8)))
ggplot(data=preds,
       aes(x=hp,y=estimate,color=as.factor(cyl))) + 
  geom_line(size=1) + 
  labs(y='MPG',x='HP',color='Cyl') + 
  theme_bw()
```

---

\textbf{Continuous} $\times$ \textbf{Categorical}

\small
$$\text{MPG}_i=(36)+(-0.1)\text{H}_i+(-15)\text{C}_i^{\text{6}}+(-18)\text{C}^8+(0.1)(\text{H}_i\times\text{C}_i^6)+(0.1)(\text{H}_i\times\text{C}_i^8)$$

* $\beta_1$: Across all cars at the reference level for Cylinder (Cylinder=4), a 1 unit increase in Horsepower is associated with -0.1 lower MPG. 
* $\beta_2$: Across all cars at the reference level for Horsepower (Horsepower=0), Cylinder=6 is associated with -15 lower MPG. 
* $\beta_3$: Across all cars at the reference level for Horsepower (Horsepower=0), Cylinder=8 is associated with -18 lower MPG. 
* $\beta_4$: The association between Horsepower and MPG for Cylinder=4 (-0.1) is 0.1 higher for Cylinder=6. The association between Cylinder=6 and MPG for Horsepower=0 (-15) is 0.1 higher for every 1 unit increase in Horsepower. 
* $\beta_5$: The association between Horsepower and MPG for Cylinder=4 (-0.1) is 0.1 higher for Cylinder=8. The association between Cylinder=8 and MPG for Horsepower=0 (-18) is 0.1 higher for every 1 unit increase in Horsepower. 

---

\frametitle{Interactions Review}

\textbf{Continuous} $\times$ \textbf{Continuous}

$$\text{MPG}_i=\beta_0+\beta_1\text{H}_i+\beta_2\text{W}_i+\beta_3(\text{H}_i\times\text{W}_i)$$

---

\textbf{Continuous} $\times$ \textbf{Continuous}

\small
$$\text{MPG}_i=\beta_0+\beta_1\text{H}_i+\beta_2\text{W}_i+\beta_3(\text{H}_i\times\text{W}_i)$$

\tiny
```{r, echo=F} 
mod <- lm(mpg~hp*wt, data=mtcars)
summary(mod)
```

---

\textbf{Continuous} $\times$ \textbf{Continuous}

\small
$$\text{MPG}_i=(49.8)+(-0.12)\text{H}_i+(-8.2)\text{W}_i+(0.02)(\text{H}_i\times\text{W}_i)$$

\tiny
```{r, echo=F} 
mod <- lm(mpg~hp*wt, data=mtcars)
summary(mod)
```

---

\textbf{Continuous} $\times$ \textbf{Continuous}

\small
$$\text{MPG}_i=(49.8)+(-0.12)\text{H}_i+(-8.2)\text{W}_i+(0.02)(\text{H}_i\times\text{W}_i)$$

```{r, echo=F} 
preds <- predictions(mod, 
                     condition = c('hp','wt'), 
                     newdata = datagrid(hp=50:350,wt=1:5))
ggplot(data=preds,
       aes(x=hp,y=estimate,color=as.factor(wt))) + 
  geom_line(size=1) + 
  labs(y='MPG',x='HP',color='Weight') + 
  theme_bw()
```

---

\textbf{Continuous} $\times$ \textbf{Continuous}

\small
$$\text{MPG}_i=(49.8)+(-0.12)\text{H}_i+(-8.2)\text{W}_i+(0.02)(\text{H}_i\times\text{W}_i)$$

* $\beta_1$: Across all cars at the reference level for Weight (Weight=0), a 1 unit increase in Horsepower is associated with -0.12 lower MPG. 

* $\beta_2$: Across all cars at the reference level for Horsepower (Horsepower=0), a 1 unit increase in Weight is associated with -8.2 lower MPG. 

* $\beta_3$: The association between Horsepower and MPG for Weight=0 (-0.12) is 0.02 higher for every 1 unit increase in Weight. The association between Weight and MPG for Horsepower=0 (-8.2) is 0.02 higher for every 1 unit increase in Horsepower.

---

\frametitle{Two-way fixed effects}

>- Two-way fixed effects and brief introduction to difference-in-differences.

\tiny
```{r, echo=F}
data("Fatalities")
Fatalities <- as.tibble(Fatalities)
Fatalities <- Fatalities %>% 
  mutate(rate = fatal / pop * 100000)
```

\tiny
```{r, echo=T}
summary(feols(rate~beertax | state+year, data=Fatalities))
```

---

\frametitle{Two-way fixed effects}

\begin{center}
\includegraphics[width = 0.8\textwidth]{figures/did.png}
\end{center}

---

\frametitle{Review: Fixed effects}

>- \textbf{Panel data:} A data set where we observe the same units (e.g., individuals, firms, countries, schools) over more than one time period. 


>- \textbf{Fixed effects model:} Basically the same thing as controlling for a categorical variable, but the "category" is units.
>   - Intercept shift by unit.
>   - Isolates our target comparison to be based on average _within-unit_ variation (i.e., variation over time). 
>   - From a causal perspective, eliminates all \textbf{time-invariant} confounders.

---

\frametitle{Review: Fixed effects}

Notation:

$$\text{y}_{i,t}=\beta_0+{\beta_1}\text{X}_{i,t}+\beta_2Z+\alpha_i$$

Estimation:

* Fixed effects estimators (modified OLS): \textbf{feols().}

```{r, echo=T, eval=F}
feols(rate ~ beertax | state, data=Fatalities)
```

---

\frametitle{Review: Fixed effects}

\textbf{Caveats:}

* \textbf{(1)} Unit-level fixed effects cannot address \textbf{time-variant} unobserved confounding. 

* \textbf{(2)} Unit-level fixed effects can lead to substantial bias if there is strong reverse causality. 

* \textbf{(3)} Including fixed effects isn't just a "better" way to estimate the same comparison; it _changes_ the comparison.



