---
title: "SOC-5811 Week 11: Panel Data"
author: "Nick Graetz"
date: "11/10/2025"
output: 
    beamer_presentation:
        incremental: false
        includes: 
            in_header: Minnesota.tex
---

```{r setup, include=FALSE}
library(AER)
library(data.table)
library(knitr)
library(kableExtra)
library(arrow)
library(haven)
library(ggplot2)
library(marginaleffects)
library(tidyverse)
library(fixest)
knitr::opts_chunk$set(
 fig.width = 6,
 fig.height = 4,
 warning = FALSE, 
 message = FALSE,
 out.width="80%"
)
dropbox <- 'C:/Users/ncgra/Dropbox/'
fig_dir <- paste0(dropbox, 'Minnesota/repos/soc5811/figures/')
```

---

\frametitle{Review: Interactions}

$$\text{PolViews}_i=\beta_0+\beta_1\text{Religion}_i+\beta_2\text{Age}_i$$

---

\frametitle{Review: Interactions}

$$\text{PolViews}_i=\beta_0+\beta_1\text{Religion}_i+\beta_2\text{Age}_i+\beta_3(\text{Religion}_i\times\text{Age}_i)$$

---

\frametitle{Review: Interactions}

$$\text{PolViews}_i=\beta_0+\boxed{\beta_1}\text{Religion}_i+\beta_2\text{Age}_i+\beta_3(\text{Religion}_i\times\text{Age}_i)$$

---

\frametitle{Review}

>- What do "fixed effects" do in linear regression models fit to panel data?

$$y_{i,t}=\beta_0+\beta_1X_{i,t}+\alpha_i$$

>- Basically the same thing as controlling for a categorical variable, but the "category" is units.
>   - Intercept shift by unit.
>   - Isolates our target comparison to be based on average _within-unit_ variation (i.e., variation over time). 
>   - From a causal perspective, eliminates all \textbf{time-invariant} confounders.

---

\frametitle{Review}

* \textbf{(1)} Unit-level fixed effects cannot address \textbf{time-variant} unobserved confounding. 

* \textbf{(2)} Unit-level fixed effects can lead to substantial bias if there is strong reverse causality. 

* \textbf{(3)} Including fixed effects isn't just a "better" way to estimate the same comparison; it _changes_ the comparison.

---

\frametitle{Review}

>- When do I want to estimate coefficients using \textbf{between-unit} differences vs. \textbf{within-unit} differences? 

>- If you care about estimating \textbf{causal effects} (or something you want to generally interpret that way), you probably want to use within-unit differences.

>- If you care about estimating \textbf{descriptive disparities}, you probably want to use between-unit differences. 

---

\frametitle{Between vs. within}

$$i=\text{hospital}$$
$$t=\text{year}$$

$$\text{Physicians}_{i,t}=\beta_0+\beta_1\text{SafetyNet}_{i,t}+\beta_2Z$$

$$\text{Physicians}_{i,t}=\beta_0+\beta_1\text{SafetyNet}_{i,t}+\beta_2Z+\alpha_i$$

---

\frametitle{Between vs. within}

$$\text{Physicians}_{i,t}=\beta_0+\boxed{\beta_1}\text{SafetyNet}_{i,t}+\beta_2Z$$

$$\text{Physicians}_{i,t}=\beta_0+\boxed{\beta_1}\text{SafetyNet}_{i,t}+\beta_2Z+\alpha_i$$

---

\frametitle{Between vs. within}

>- \textbf{Research question:} Do safety net hospitals have different average physicians per patient, controlling for $Z$? 
>- $\beta_1$: Across all hospitals, safety net hospitals have $\beta_1$ more/fewer physicians per capita.
>- Descriptive comparison with implications for stratification/inequality. 

$$\text{Physicians}_{i,t}=\beta_0+\boxed{\beta_1}\text{SafetyNet}_{i,t}+\beta_2Z$$

---

\frametitle{Between vs. within}

>- \textbf{Research question:} As the patient population changes within hospitals, does average physicians per patient change? 
>- Causal comparison with implications for the effect of patient population on work force within hospitals. When hospitals become safety net hospitals, do they see reductions in work force? 

$$\text{Physicians}_{i,t}=\beta_0+\boxed{\beta_1}\text{Pop}_{i,t}+\beta_2Z+\alpha_i$$

---

\frametitle{Between vs. within}

* \textbf{Research question:} Do safety net hospitals have different average physicians per patient, controlling for $Z$? 

* \textbf{Research question:} As the patient population changes within hospitals, does average physicians per patient change?  

* \textbf{Both answer interesting but different research questions.}

---

\frametitle{Estimation}

How to fit fixed effects models: 

>- OLS estimator: Add unit as a categorical variable in lm().

>- OLS estimator: Manually demean and then lm() with demeaned data.

>- Fixed effects estimators (modified OLS): \textbf{feols().}

\tiny
```{r, echo=F}
data("Fatalities")
Fatalities <- as.tibble(Fatalities)
Fatalities <- Fatalities %>% 
  mutate(rate = fatal / pop * 100000)
Fatalities <- Fatalities %>%
  group_by(state) %>% 
  mutate(rate_demeaned = rate - mean(rate),
         tax_demeaned = beertax - mean(beertax))
```

---

\frametitle{Estimation}

\tiny
```{r, echo=T}
summary(lm(rate ~ beertax + state, data=Fatalities))
```

---

\frametitle{Estimation}

\tiny
```{r, echo=T}
summary(lm(rate_demeaned ~ tax_demeaned, data=Fatalities))
```

---

\small
```{r, echo=T}
summary(feols(rate ~ beertax | state, data=Fatalities))
```

---

\frametitle{Estimation}

How to fit fixed effects models: 

>- OLS estimator: Add unit as a categorical variable in lm().
>   - Very slow/clunky if you have thousands of units. 

>- OLS estimator: Manually demean and then lm() with demeaned data.
>   - Very complex to do correctly by hand if you have many variables in your model. 

>- Fixed effects estimators (modified OLS): \textbf{feols().}
>   - Very fast and efficient; the tradeoff is that you don't estimate standard errors on your fixed effects themselves, but typically this does not matter for your target comparison. 

---

\frametitle{Uncertainty}

>- It is tricky to get the standard errors correct with panel data. Why?

>- Remember that in the conventional regression model, we assume errors are \textbf{independent.} In sampling terms, we assume every sample from the population model is independent of any other sample.

>- Think about flipping a coin 100 times. When we estimate the standard error of the random variable $X$ = number of heads in 100 flips, we assume each flip is unrelated to (or independent from) the other flips.

>- Solution: \textbf{clustered standard errors.}

---

\frametitle{Uncertainty}

\small
\textbf{$\beta_1$ from our basic OLS model with unit as a categorical variable:}

```{r, echo=F}
mod1 <- summary(lm(rate ~ beertax + state, data=Fatalities))
mod1[['coefficients']][2,]
```
\textbf{$\beta_1$ from our fixed effects model:}

```{r, echo=F}
mod2 <- summary(feols(rate ~ beertax | state, data=Fatalities))
summary(mod2)
```

---

\frametitle{Review}

Pick up here on Wednesday.

---

\frametitle{Two-way fixed effects}

>- Two-way fixed effects and brief introduction to difference-in-differences.

