---
title: "SOC-5811 Week 14: Transformations"
author: "Nick Graetz"
date: "12/1/2025"
output: 
    beamer_presentation:
        incremental: false
        includes: 
            in_header: Minnesota.tex
---

```{r setup, include=FALSE}
set.seed(1234)
library(AER)
library(data.table)
library(knitr)
library(kableExtra)
library(arrow)
library(haven)
library(ggplot2)
library(marginaleffects)
library(tidyverse)
library(fixest)
library(cowplot)
knitr::opts_chunk$set(
 fig.width = 6,
 fig.height = 4,
 warning = FALSE, 
 message = FALSE,
 out.width="80%"
)
dropbox <- 'C:/Users/ncgra/Dropbox/'
fig_dir <- paste0(dropbox, 'Minnesota/repos/soc5811/figures/')
```

---

\frametitle{Transformations}

>- How do we relax our assumptions around additivity and linearity?
>   - Additivity: add interactions.
>   - Linearity: transform variables.

>- What does "relax" mean in this context? 
>   - Rule of thumb: Inferences made with simpler models rely on _more_ assumptions.

---

\frametitle{Transformations}

$$\text{Happiness}_i=\beta_0+\beta_1\text{Income}_i$$

Say I want to use this model to infer whether people with 150k are happier, on average, than people with 50k. 

\textbf{What assumption does this inference rely on?}

---

\frametitle{Transformations}

\textbf{What are "transformations?"}

>- We aren't talking about _shifting_ or _rescaling_ predictor variables, which changes our interpretation of the coefficients on those variables.

>- We are talking about changing the _functional relationship_ between a predictor variable and the outcome. 

---

\frametitle{Transformations}

>- Log-transforming

>- Polynomials

>- Splines

---

\frametitle{Transformations}

$$y_i=\beta_0+\beta_1X_i$$

```{r, echo=F}
x <- seq(0, 10, length.out = 100)
y <- 2 * x^2 - 5 * x + 3 + rnorm(100, 0, 2)
d <- data.table(x=x,y=y)
mod <- lm(y~x, data=d)
d[, y_hat := predict(mod)]
ggplot(data=d,
       aes(x=x,
           y=y)) + 
  geom_point(shape=21,size=2,fill='lightgrey') + 
  geom_line(aes(y=y_hat),size=1) +
  theme_bw()
```

---

\frametitle{Transformations}

$$y_i=\beta_0+\beta_1X_i+\beta_2X_i^2$$

```{r, echo=F}
x <- seq(1, 10, length.out = 100)
# y <- rexp(n = 1000, rate = 20 * x + 3 + rnorm(1000, 0, 3))
y <- 2 * x^2 - 5 * x + 3 + rnorm(100, 0, 2)
d <- data.table(x=x,y=y)
d[, x2 := x^2]
mod <- lm(y~x+x2, data=d)
d[, y_hat := predict(mod)]
ggplot(data=d,
       aes(x=x,
           y=y)) + 
  geom_point(shape=21,size=2,fill='lightgrey') + 
  geom_line(aes(y=y_hat),size=1) +
  theme_bw()
```

---

\frametitle{Transformations}

$$y_i=\beta_0+\beta_1\log(X_i)$$

```{r, echo=F}
x <- seq(1, 10, length.out = 100)
y <- 10 * log(x) + 3 + rnorm(100, 0, 2)
d <- data.table(x=x,y=y)
d[, log_x := log(x)]
mod <- lm(y~log_x, data=d)
d[, y_hat := predict(mod)]
ggplot(data=d,
       aes(x=x,
           y=y)) + 
  geom_point(shape=21,size=2,fill='lightgrey') + 
  geom_line(aes(y=y_hat),size=1) +
  theme_bw()
```

---

\frametitle{Transformations}

$$y_i=\beta_0+\beta_1\log(X_i)$$
\tiny
```{r, echo=T}
summary(mod)
```
$$\beta_1=9.6$$

---

\frametitle{Transformations}

$$y_i=\beta_0+\beta_1\log(X_i)$$

$$\beta_1=9.6$$

Interpretation:

>- A one-unit increase in ln(X) is associated with a 9.6 increase in Y. 
>- A 1% increase in X is associated with a .096 increase in Y. 

---

\frametitle{Transformations}

```{r, echo=F}
x <- seq(1, 10, length.out = 100)
y <- rlnorm(n = 1000, meanlog = x, sdlog = 0.3)
y[y==0] <- 0
d <- data.table(x=x,y=y)
hist(y)
```

---

\frametitle{Transformations}

$$y_i=\beta_0+\beta_1(X_i)$$

```{r, echo=F}
d[, log_x := log(x)]
d[, log_y := log(y)]
mod <- lm(y~x, data=d)
d[, y_hat := predict(mod)]
ggplot(data=d,
       aes(x=x,
           y=y)) + 
  geom_point(shape=21,size=2,fill='lightgrey') + 
  geom_line(aes(y=y_hat),size=1) +
  theme_bw()
```

---

\frametitle{Transformations}

$$\log(y_i)=\beta_0+\beta_1(X_i)$$

```{r, echo=F}
d[, log_x := log(x)]
d[, log_y := log(y)]
mod <- lm(log_y~x, data=d)
d[, y_hat := exp(predict(mod))]
ggplot(data=d,
       aes(x=x,
           y=y)) + 
  geom_point(shape=21,size=2,fill='lightgrey') + 
  geom_line(aes(y=y_hat),size=1) +
  theme_bw()
```

---

\frametitle{Transformations}

\textbf{Why would we do this?} 

>- Avoid negative predictions for an continuous outcome variable that is bounded at 0. 

>- If we think the relationship between X and Y is multiplicative rather than additive. 

---

\frametitle{Transformations}

>- \textbf{Only the dependent/response variable is log-transformed.} Exponentiate the coefficient. This gives the multiplicative factor for every one-unit increase in the independent variable. 

>- \textbf{Only independent/predictor variable(s) is log-transformed.} Divide the coefficient by 100. This tells us that a 1% increase in the independent variable increases (or decreases) the dependent variable by (coefficient/100) units.

---

\frametitle{Transformations}

>- \textbf{Both dependent/response variable and independent/predictor variable(s) are log-transformed.} Interpret the coefficient as the percent increase in the dependent variable for every 1% increase in the independent variable.

---

\frametitle{Transformations}

>- Log-transforming

>- Polynomials

>- \textbf{Splines}

---

\frametitle{Transformations}

```{r, echo=F}
library(lspline)
x <- seq(from = 0, to = 6, by = .025)
y <- sin(2*x) + x -.1*x^2 + 2 + rnorm(length(x), sd = .3)
d <- data.table(x,y)
```

```{r, echo=T}
mod <- lm(y~x, data=d)
```

```{r, echo=F}
preds_ols <- predictions(mod,by='x')
ggplot(data=d,
       aes(x=x,y=y)) + 
  geom_point() + 
  geom_line(data=preds_ols,aes(x=x,y=estimate), color='black', size=1) + 
  geom_ribbon(data=preds_ols,aes(x=x,y=estimate,ymin=conf.low,ymax=conf.high), fill='black', alpha=0.3) +
  theme_bw()
```

---

\frametitle{Transformations}

```{r, echo=T}
mod <- lm(y~lspline(x,knots=3), data=d)
```

```{r, echo=F}
preds <- predictions(mod,by='x')
ggplot(data=d,
       aes(x=x,y=y)) + 
  geom_point() + 
  geom_line(data=preds_ols,aes(x=x,y=estimate), color='black', size=1) +
  geom_ribbon(data=preds_ols,aes(x=x,y=estimate,ymin=conf.low,ymax=conf.high), fill='black', alpha=0.3) +
  geom_line(data=preds,aes(x=x,y=estimate), color='red', size=1) +
  geom_ribbon(data=preds,aes(x=x,y=estimate,ymin=conf.low,ymax=conf.high), fill='red', alpha=0.3) +
  theme_bw()
```

---

\frametitle{Transformations}

\tiny
```{r, echo=F}
summary(mod)
```

---

\frametitle{Transformations}

```{r, echo=T}
mod <- lm(y~lspline(x,knots=c(1,2,4)), data=d)
```

```{r, echo=F}
preds <- predictions(mod,by='x')
ggplot(data=d,
       aes(x=x,y=y)) + 
  geom_point() + 
  geom_line(data=preds_ols,aes(x=x,y=estimate), color='black', size=1) +
  geom_ribbon(data=preds_ols,aes(x=x,y=estimate,ymin=conf.low,ymax=conf.high), fill='black', alpha=0.3) +
  geom_line(data=preds,aes(x=x,y=estimate), color='red', size=1) +
  geom_ribbon(data=preds,aes(x=x,y=estimate,ymin=conf.low,ymax=conf.high), fill='red', alpha=0.3) +
  theme_bw()
```

---

\frametitle{Transformations}

```{r, echo=T}
mod <- lm(y~pspline(x,degree=1), data=d)
```

```{r, echo=F}
preds_linear <- predictions(mod,by='x')
ggplot(data=d,
       aes(x=x,y=y)) + 
  geom_point() + 
  geom_line(data=preds_ols,aes(x=x,y=estimate), color='black', size=1) +
  geom_ribbon(data=preds_ols,aes(x=x,y=estimate,ymin=conf.low,ymax=conf.high), fill='black', alpha=0.3) +
  geom_line(data=preds_linear,aes(x=x,y=estimate), color='red', size=1) +
  geom_ribbon(data=preds_linear,aes(x=x,y=estimate,ymin=conf.low,ymax=conf.high), fill='red', alpha=0.3) +
  theme_bw()
```

---

\frametitle{Transformations}

```{r, echo=T}
mod <- lm(y~pspline(x,degree=3), data=d)
```

```{r, echo=F}
preds_cubic <- predictions(mod,by='x')
ggplot(data=d,
       aes(x=x,y=y)) + 
  geom_point() + 
  geom_line(data=preds_ols,aes(x=x,y=estimate), color='black', size=1) +
  geom_ribbon(data=preds_ols,aes(x=x,y=estimate,ymin=conf.low,ymax=conf.high), fill='black', alpha=0.3) +
  geom_line(data=preds_cubic,aes(x=x,y=estimate), color='red', size=1) +
  geom_ribbon(data=preds_cubic,aes(x=x,y=estimate,ymin=conf.low,ymax=conf.high), fill='red', alpha=0.3) +
  theme_bw()
```

---

\frametitle{Transformations}

Penalized cubic splines

```{r, echo=F}
ggplot(data=d,
       aes(x=x,y=y)) + 
  geom_point() + 
  geom_line(data=preds_ols,aes(x=x,y=estimate), color='black', size=1) +
  geom_ribbon(data=preds_ols,aes(x=x,y=estimate,ymin=conf.low,ymax=conf.high), fill='black', alpha=0.3) +
  geom_line(data=preds_cubic,aes(x=x,y=estimate), color='red', size=1) +
  geom_ribbon(data=preds_cubic,aes(x=x,y=estimate,ymin=conf.low,ymax=conf.high), fill='red', alpha=0.3) +
  theme_bw()
```

---

\frametitle{Transformations}

Which model do we like? Linear or penalized cubic spline?

```{r, echo=F}
ggplot(data=d,
       aes(x=x,y=y)) + 
  geom_point() + 
  geom_line(data=preds_ols,aes(x=x,y=estimate), color='black', size=1) +
  geom_ribbon(data=preds_ols,aes(x=x,y=estimate,ymin=conf.low,ymax=conf.high), fill='black', alpha=0.3) +
  geom_line(data=preds_cubic,aes(x=x,y=estimate), color='red', size=1) +
  geom_ribbon(data=preds_cubic,aes(x=x,y=estimate,ymin=conf.low,ymax=conf.high), fill='red', alpha=0.3) +
  theme_bw()
```

---

\frametitle{Transformations}

\tiny
```{r, echo=F}
summary(mod)
```

---

\frametitle{Transformations}

```{r, echo=F}
x <- seq(from = 0, to = 6, by = .025)
y <- sin(2*x) + x -.1*x^2 + 2 + rnorm(length(x), sd = 2)
d <- data.table(x,y)
mod <- lm(y~x, data=d)
preds_linear <- predictions(mod,by='x')
mod <- lm(y~pspline(x,degree=3), data=d)
preds_cubic <- predictions(mod,by='x')
ggplot(data=d,
       aes(x=x,y=y)) + 
  geom_point() + 
  geom_line(data=preds_ols,aes(x=x,y=estimate), color='black', size=1) +
  geom_ribbon(data=preds_ols,aes(x=x,y=estimate,ymin=conf.low,ymax=conf.high), fill='black', alpha=0.3) +
  geom_line(data=preds_cubic,aes(x=x,y=estimate), color='red', size=1) +
  geom_ribbon(data=preds_cubic,aes(x=x,y=estimate,ymin=conf.low,ymax=conf.high), fill='red', alpha=0.3) +
  theme_bw()
```

---

\frametitle{Transformations}

```{r, echo=F}
x <- 1:100
y <- 3*x + rnorm(100,mean=0,sd=25)
d <- data.table(x,y)
mod <- lm(y~x, data=d)
preds_linear <- predictions(mod,by='x')
mod <- lm(y~pspline(x,degree=3), data=d)
preds_cubic <- predictions(mod,by='x')
ggplot(data=d,
       aes(x=x,y=y)) + 
  geom_point() + 
  geom_line(data=preds_linear,aes(x=x,y=estimate), color='black', size=1) +
  geom_ribbon(data=preds_linear,aes(x=x,y=estimate,ymin=conf.low,ymax=conf.high), fill='black', alpha=0.3) +
  geom_line(data=preds_cubic,aes(x=x,y=estimate), color='red', size=1) +
  geom_ribbon(data=preds_cubic,aes(x=x,y=estimate,ymin=conf.low,ymax=conf.high), fill='red', alpha=0.3) +
  theme_bw()
```

---

\frametitle{Transformations}

>- Splines vs. polynomials

```{r, echo=F}
x <- seq(1, 10, length.out = 100)
y <- 2 * x^2 - 5 * x + 3 + rnorm(100, 0, 5)
d <- data.table(x=x,y=y)
d[, x2 := x^2]
mod <- lm(y~x+x2, data=d)
preds_poly <- predictions(mod,by='x')
mod <- lm(y~pspline(x,degree=3), data=d)
preds_cubic <- predictions(mod,by='x')
ggplot(data=d,
       aes(x=x,y=y)) + 
  geom_point() + 
  geom_line(data=preds_poly,aes(x=x,y=estimate), color='black', size=1) +
  geom_ribbon(data=preds_poly,aes(x=x,y=estimate,ymin=conf.low,ymax=conf.high), fill='black', alpha=0.3) +
  geom_line(data=preds_cubic,aes(x=x,y=estimate), color='red', size=1) +
  geom_ribbon(data=preds_cubic,aes(x=x,y=estimate,ymin=conf.low,ymax=conf.high), fill='red', alpha=0.3) +
  theme_bw()
```

---

\frametitle{Transformations}

>- \textbf{Quiz:} I'm interesting in the effect of $X$ (i.e., $\beta_1$), and $Z$ is a potential confounder. Inference on $\beta_1$ requires more assumptions in which of these models?

$$y_i=\beta_0+\beta_1X_i+\beta_2Z_i$$

$$y_i=\beta_0+\beta_1X_i+\beta_2S(Z_i)$$

>- \textbf{Rule of thumb:} Inferences made with simpler models rely on _more_ assumptions.

---

\frametitle{Transformations}

>- \textbf{Quiz:} If we want to relax our assumptions of additivity and linearity as much as possible, why not just use splines for all continuous variables and include all interactions?

>- There are models out there besides the conventional linear model that basically do this!
>   - How do they handle the complexity vs. overfitting tradeoff? Cross-validation.

---

\frametitle{Transformations}

\textbf{Why do we sometimes prefer simpler models?}

>- We want parameter estimates that are easy to interpret.
>   - This is the weakest reason, because we can always just use conditional/marginal predictions for our inferences even if the model is very complicated in the parameters. 
>   - But it is also the reason that has historically driven the use of simpler models; it is very easy to estimate the linear regression coefficient.

>- We care about summarizing with average effects. 

>- We can demonstrate that our assumptions (e.g., linearity) do not substantively impact our conclusions. 
