---
title: "SOC-5811 Week 3: Intro to R and regression"
author: "Nick Graetz"
date: "9/17/2025"
output: 
    beamer_presentation:
        incremental: false
        includes: 
            in_header: Minnesota.tex
---

```{r setup, include=FALSE}
## Libraries
library(tidyverse)
library(data.table)
library(ggplot2)
library(haven)
library(formatR)
library(marginaleffects)
# knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 20), tidy = TRUE)
knitr::opts_chunk$set(
 fig.width = 6,
 fig.height = 4
)
```

---

\frametitle{Load data}

>- R is an \textbf{object-oriented} programming language. 
>- We assign objects with the \textbf{<-} operator.
>- We apply functions to objects: \textbf{function(object)}.
>- All objects have a \textbf{class}.
>- All functions take inputs of certain classes and return outputs of certain classes.

>- \textbf{Scripts are run in computing environments.}

---

\frametitle{Set up filepaths}

* Assign my filepath to an object called "dropbox"
* Test different basic R functions
    
\small
```{r, echo=TRUE, error=TRUE}
dropbox <- 'C:/Users/ncgra/Dropbox/'
class(dropbox)
length(dropbox)
nchar(dropbox)
```
---

\frametitle{Set up filepaths}
    
\small
```{r, echo=TRUE, error=TRUE}
sum(dropbox)
```

---

\frametitle{Set up filepaths}

* I'm going to set up a few filepaths.

\small
```{r, echo=TRUE}
fig_dir <- paste0(dropbox,
                  'Minnesota/repos/soc5811/figures/')
data_dir <- paste0(dropbox,
                   'Minnesota/repos/soc5811/data/')
```

---

\frametitle{Load data}

* We are going to look at population and housing data from the 2000/2010 Census.

\small
```{r, echo=TRUE}
census <- read_dta(paste0(data_dir,
                          'state_pophouse.dta'))
```

---

\frametitle{Load data}

\tiny
```{r, echo=TRUE}
class(census)
dim(census)
names(census)
head(census)
```

---

\frametitle{Examine data}

We can use different functions like select() and slice() to look at specific rows and columns:

\tiny
```{r, echo=TRUE}
census %>% 
  select(state,a00aa2000) %>%
  slice(1:5)
```

---

\frametitle{Examine data}

We can use other packages like "data.table" with different functions:

\tiny
```{r, echo=TRUE}
census <- as.data.table(census)
class(census)
census[1:5, c('state','a00aa2000')]
```
---

\frametitle{Examine data}

ggplot and the Grammar of Graphics

https://ggplot2-book.org/introduction.html 

\small
```{r, echo=TRUE}
plot <- ggplot(data=census, 
               aes(x=pctpop, 
                   y=pcthouse)) + 
  geom_point()
```

---

\frametitle{Examine data}

\small
```{r, echo=FALSE}
print(plot)
```

---

\frametitle{Examine data}

\small
```{r, echo=TRUE}
plot <- ggplot(data=census, 
               aes(x=pctpop, 
                   y=pcthouse)) + 
  geom_point() + 
  labs(x='Percent pop growth 2000-2010',
       y='Percent housing growth 2000-2010') + 
  theme_bw()
```

---

\frametitle{Examine data}

\small
```{r, echo=FALSE}
print(plot)
```

---

\frametitle{Examine data}

\small
```{r, echo=TRUE}
plot <- ggplot(data=census, 
               aes(x=pctpop, 
                   y=pcthouse, 
                   label=state)) + 
  geom_label(size=3) + 
  labs(x='Percent pop growth 2000-2010',
       y='Percent housing growth 2000-2010') + 
  theme_bw()
```

---

\frametitle{Examine data}

\small
```{r, echo=FALSE}
print(plot)
```

---

\frametitle{Population regression functions}

Let's think about creating a model for housing growth:

$$pcthouse = f(pctpop)$$

---

\frametitle{Population regression functions}

Let's think about creating a model for housing growth:

$$pcthouse = f(pctpop)$$

$$pcthouse_i=\beta_0+\beta_1{pctpop_i}+\epsilon_i$$

---

\frametitle{Population regression functions}

Fitting a linear regression with data:

\small
```{r, echo=TRUE}
model <- lm(pcthouse~pctpop, 
            data=census)
```

\tiny
```{r, echo=TRUE}
summary(model)
```

---

\frametitle{Making predictions}

\tiny
```{r, echo=TRUE}
census <- census %>% 
  mutate(pcthouse_pred=predict(model))
head(census)
```

---

\frametitle{Making predictions}

\tiny
```{r, echo=FALSE}
ggplot(data=census,
       aes(x=pctpop,
           y=pcthouse)) + 
  geom_line(aes(x=pctpop,y=pcthouse_pred)) + 
  geom_point() + 
  labs(x='Percent pop growth 2000-2010',
       y='Percent housing growth 2000-2010') + 
  theme_bw()
ggsave(paste0(fig_dir,'/week3_scatter.png'))
```

---

\frametitle{Making predictions}

\tiny
```{r, echo=FALSE}
census <- census %>%
  mutate(residual=pcthouse-pcthouse_pred)
ggplot(data=census,
       aes(x=pctpop,
           y=pcthouse)) + 
  geom_line(aes(x=pctpop,y=pcthouse_pred)) + 
  geom_linerange(data=census %>% filter(residual>=0),
                 aes(x=pctpop,ymin=pcthouse_pred,ymax=pcthouse),
                 color='blue') + 
  geom_linerange(data=census %>% filter(residual<0),
                 aes(x=pctpop,ymin=pcthouse,ymax=pcthouse_pred),
                 color='red') + 
  geom_point() + 
  labs(x='Percent pop growth 2000-2010',
       y='Percent housing growth 2000-2010') + 
  theme_bw()
```

---

\frametitle{Building models}

\small
\begin{columns}[onlytextwidth,T]
  \begin{column}{.4\linewidth}
  What can we do with this model?
    \begin{enumerate}
      \item Generalizing from sample to population.
      \item Measurement.
      \item Forecasting.
      \item Causal inference.
    \end{enumerate}
  \end{column}
  \begin{column}{.6\linewidth}
  \includegraphics[width=\textwidth,height=0.80\textheight,keepaspectratio]{figures/week3_scatter}
  \end{column}
\end{columns}

---

\frametitle{Building models}

\small
\begin{columns}[onlytextwidth,T]
  \begin{column}{.4\linewidth}
  What can we do with this model?
    \begin{enumerate}
      \item \textbf{Generalizing from sample to population:} Is this coefficient the same one I would estimate with the entire population?
    \end{enumerate}
  \end{column}
  \begin{column}{.6\linewidth}
  \includegraphics[width=\textwidth,height=0.80\textheight,keepaspectratio]{figures/week3_scatter}
  \end{column}
\end{columns}

---

\frametitle{Building models}

\small
\begin{columns}[onlytextwidth,T]
  \begin{column}{.4\linewidth}
  What can we do with this model?
    \begin{enumerate}
      \item Generalizing from sample to population
      \item \textbf{Measurement:} Can I generalize to all types of housing growth?
    \end{enumerate}
  \end{column}
  \begin{column}{.6\linewidth}
  \includegraphics[width=\textwidth,height=0.80\textheight,keepaspectratio]{figures/week3_scatter}
  \end{column}
\end{columns}

---

\frametitle{Building models}

\small
\begin{columns}[onlytextwidth,T]
  \begin{column}{.4\linewidth}
  What can we do with this model?
    \begin{enumerate}
      \item Generalizing from sample to population
      \item Measurement
      \item \textbf{Forecasting:} Can I use this model to predict out-of-sample?
    \end{enumerate}
  \end{column}
  \begin{column}{.6\linewidth}
  \includegraphics[width=\textwidth,height=0.80\textheight,keepaspectratio]{figures/week3_scatter}
  \end{column}
\end{columns}

---

\frametitle{Building models}

\small
\begin{columns}[onlytextwidth,T]
  \begin{column}{.4\linewidth}
  What can we do with this model?
    \begin{enumerate}
      \item Generalizing from sample to population
      \item Measurement
      \item Forecasting
      \item \textbf{Causal inference:} Can I say pop growth \textbf{causes} housing growth?
    \end{enumerate}
  \end{column}
  \begin{column}{.6\linewidth}
  \includegraphics[width=\textwidth,height=0.80\textheight,keepaspectratio]{figures/week3_scatter}
  \end{column}
\end{columns}

---

\frametitle{Building models}

\tiny
```{r, echo=FALSE}
ggplot(data=census,
       aes(x=pctpop,
           y=pcthouse)) + 
  geom_line(aes(x=pctpop,y=pcthouse_pred),color='blue') +
  geom_line(aes(x=pctpop,y=pcthouse),color='red') + 
  geom_point() + 
  labs(x='Percent pop growth 2000-2010',
       y='Percent housing growth 2000-2010') + 
  theme_bw()
```

